### 2016-11-23 latex/article.tex
Added latex file

\documentclass[11pt,a4paper]{article}
\linespread{1.25}
\usepackage[utf8]{inputenc} % Norwegian letters
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{float}  % Used for minipage and stuff.
\usepackage{wrapfig} % Wrap text around figure wrapfig [tab]
\usepackage{graphicx}
\usepackage{enumerate} % Use e.g. \begin{enumerate}[a)]
\usepackage[font={small, it}]{caption} % captions on figures and tables
\usepackage{subcaption}
\usepackage[toc,page]{appendix} % Page make Appendices title, toc fix table of content 
%\usepackage{todonotes} % Notes. Use \todo{"text"}. Comment out \listoftodos
\usepackage{xargs}                      % Use more than one optional parameter in a new commands
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.
% 
\usepackage[colorinlistoftodos,prependcaption,textsize=footnotesize]{todonotes}
\newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
\newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
\newcommandx{\info}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}}
\newcommandx{\improvement}[2][1=]{\todo[linecolor=Plum,backgroundcolor=Plum!25,bordercolor=Plum,#1]{#2}}
\newcommandx{\thiswillnotshow}[2][1=]{\todo[disable,#1]{#2}}
\usepackage{microtype} % Improves spacing. Include AFTER fonts
\usepackage{hyperref} % Use \autoref{} and \nameref{}
\hypersetup{backref,
  colorlinks=true,
  breaklinks=true,
  %hidelinks, %uncomment to make links black
  citecolor=blue,
  linkcolor=blue,
  urlcolor=blue
}
\usepackage[all]{hypcap} % Makes hyperref jup to top of pictures and tables
%
%-------------------------------------------------------------------------------
% Page layout
%\usepackage{showframe} % Uncomment if you want the margin frames
\usepackage{fullpage}
\topmargin=-0.25in
%\evensidemargin=-0.3in
%\oddsidemargin=-0.3in
%\textwidth=6.9in
%\textheight=9.5in
\headsep=0.25in
\footskip=0.50in

%-------------------------------------------------------------------------------
% Header and footer
\usepackage{lastpage} % To be able to add last page in footer.
\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancy} % Use "fancyplain" for header in all pages
%\renewcommand{\chaptermark}[1]{ \markboth{#1}{} } % Usefull for book?
\renewcommand{\sectionmark}[1]{ \markright{\thesection\ #1}{} } % Remove formating and nr.
%\fancyhead[LE, RO]{\footnotesize\leftmark}
%\fancyhead[RO, LE]{\footnotesize\rightmark}
\lhead[]{\AuthorName}
\rhead[]{\rightmark}
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{Page\ \thepage\ of\ \protect\pageref*{LastPage}} % Page numbering for right footer
\renewcommand{\headrulewidth}{1pt} % header underlines
\renewcommand{\footrulewidth}{1pt} % footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

%-------------------------------------------------------------------------------
% Suppose to make it easier for LaTeX to place figures and tables where I want.
\setlength{\abovecaptionskip}{0pt plus 1pt minus 2pt} % Makes caption come closer to figure.
%\setcounter{totalnumber}{5}
%\renewcommand{\textfraction}{0.05}
%\renewcommand{\topfraction}{0.95}
%\renewcommand{\bottomfraction}{0.95}
%\renewcommand{\floatpagefraction}{0.35}
%
% Math short cuts for expectation, variance and covariance
\newcommand{\E}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\mk}[1]{\colorbox{yellow}{#1}}
% Commands for argmin and argmax
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
%	TITLE SECTION
%-----------------------------------------------------------------------------
\newcommand{\AuthorName}{Håvard Kvamme} % Your name

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{\
\normalfont \normalsize 
\textsc{STK9101SP} \\ [25pt] % Your university, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Project on Survival Analysis \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{\AuthorName} % Your name

\date{\normalsize\today} % Today's date or a custom date
\begin{document}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\listoftodos{}
%
\section{Introduction}
\todo[inline]{Introduce problem and covariates}

The dataset has tied events due to the rounding of events to nearest date. Out of the 307 observations, 296 occurs alone in a single date. However, this does not affect the estimation of the Kaplan-Meier.

\section{a) Simple univariate analysis}
\unsure[inline]{Does univariate mean only one group??? \\I would prefer to use many several groups.}
\unsure[inline]{Should I try to interpret the K-M plots through quantiles and stuff?? I’m pretty sure I’ve seen that somewhere. Show median survival times for different groups?}

\begin{figure}[h!tbp]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/km_beh.pdf}
        %\caption{Alb.}
        %\label{fig:cartAreas1}
    \end{subfigure}%
    \quad
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/km_ald.pdf}
        %\caption{not alb.}
        %\label{fig:cartTree1}
    \end{subfigure}
    %(or a blank line to force the subfigure onto a new line)
    \vspace{1\baselineskip}
    \caption{Survival curves for treatment and age. We have included histograms for the two covariates under the survival curves.}
    \label{fig:alb}
\end{figure}

\begin{figure}[h!tbp]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/km_kjonn.pdf}
        %\caption{Alb.}
        %\label{fig:cartAreas1}
    \end{subfigure}%
    \quad
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/km_bil.pdf}
        %\caption{not alb.}
        %\label{fig:cartTree1}
    \end{subfigure}
    %(or a blank line to force the subfigure onto a new line)
    \vspace{1\baselineskip}
    \caption{Survival curves for gender and amount of bilirubin in the blood of the test subjects. We have included histograms for the two covariates under the survival curves.}
    \label{fig:alb}
\end{figure}

\begin{figure}[h!tb]
    \begin{center}
        \includegraphics[scale=0.8]{./figures/km_alb.pdf}
    \end{center}
    \vspace{-0.8cm}
    \caption{Survival curve for amount of albumin in the blood of the test subjects. The histogram displays the distribution of albumin among the subjects.}
    \label{fig:alb}
\end{figure}

\begin{table}[h!tbp]
    \centering
    \caption{Log-rank tests on the individual covariates.}
    \label{tab:log_rank_indiv}
    \begin{tabular}{rrrr}
        \hline
        Covariate & deg. & Chisq. & $p$ \\ 
        \hline
        Treatment &  1 &  $0   $ & $0.853     $ \\
        Age       &  3 &  $21  $ & $1.07e-4   $ \\ 
        Gender    &  1 &  $3.5 $ & $0.0613    $ \\ 
        Bilirubin &  4 &  $182 $ & $0         $ \\ 
        Albumin   &  3 &  $68.7$ & $7.99e-15$ \\ 
        \hline
    \end{tabular}
\end{table}

\section{b) Univariate regression}
\info[inline]{For categorical covariates, cox scores and log-rank are equivalent, if not ties. What with ties??}
\unsure[inline]{How does cox handle tied data?}
\info[inline]{Cox can fail if $x$ is wrong and we should use log or sqrt instead. Or if the hazards are not proportional.}

\subsection{Treatment}

Here we fitted an cox regression to the treatment covariate. We see from Table~\ref{tab:log_rank_indiv} that the treatment and placebo group are not significantly different, and thus we do not expect the cox regression to be significant either. This is because the score test in cox regression, with categorical covariates, is equivalent to the log-rank test, as long as there are no ties. We do have ties in our data, but not that many.

Table~\ref{tab:cox_treatment} shows some of the results from the cox regression. First note that we display the exponentiated coefficient, as this gives the hazard ratio between the groups, as well as the 90\% confidence interval. We see that the hazard rate is close to one, and 1 is covered in the confidence interval. This indicates that there are no difference between the groups, as expected from our previous analysis.

We can see that the score test in Table~\ref{tab:cox_treatment} is the same as the log-rank test in Table~\ref{tab:log_rank_indiv}, which is as expected.

Also note that none of the tests for the proportional hazard assumption fails (zph).
So the cox model seems to be a reasonable model for the treatment covariate.

\input{tables/cox_treatment}

\subsection{Age}

\input{tables/cox_age}

In Figure~\ref{fig:cox_age_mart} we have plotted the martingale residuals from the cox regression of age. There does not seem to be a pattern in the residuals, so the functional form of the covariate seems appropriate.

\begin{figure}[h!tb]
    \begin{center}
        \includegraphics[scale=0.6]{./figures/cox_age_mart.pdf}
    \end{center}
    \vspace{-0.8cm}
    \caption{Martingale residuals for cox regression on age.}
    \label{fig:cox_age_mart}
\end{figure}

\subsection{Gender}
\input{tables/cox_gender}

\subsection{Bilirubin}
\input{tables/cox_bilirubin}

\subsection{Albumin}

\subsection{Testing model assumptions}

\begin{itemize}
    \item Can plot martingale residuals vs time (observation number) and color by group. To see if the groups are evenly spread. This check the PH assumption.
    \item Investigate martingale residuals. Figure this out!!!!
    \item Scaled Schoenfeld residuals (lec 9 p 35) to check proportionality over time.
    \item Schoenfeld residuals for prop. assumption (lec 9 p 38).
    \item Check proportionality by lec 9 p 29: Plot log nelson aalen for different groups and check if they are parallel.
    \item Practical exercise 8. Fit spline to check if log-linear effect (which is a model assumption?). How to interpret results?
    \item \url{http://www.mwsug.org/proceedings/2006/stats/MWSUG-2006-SD08.pdf}
\end{itemize}

\section{c) Multivariate regression}
\begin{itemize}
    \item See Practical exercise 7 for likelihood ratio test for significance of interactions.
\end{itemize}

\clearpage
\section{Dolphin Stuff}

\subsection{The Dataset}

All analysis was done on a directed social network of bottlenose dolphins \citep{dolphins}. The nodes are the bottlenose dolphins (genus Tursiops) of a bottlenose dolphin community living off Doubtful Sound, a fjord in New Zealand (spelled fiord in New Zealand). An edge indicates a frequent association. The dolphins were observed between 1994 and 2001.

Part of what could make this network interesting, is that we can study the social interactions between highly intelligent animal other than humans. So we could for instance compare the dolphins to different species, to investigate the impact of intelligence on social structures. However, I will do no such analysis, and I will admit that the network was chosen partly because of its size. It only contains 62 dolphins, so much of the analysis is not particularly computationally expensive.
The network is displayed in Figure~\ref{fig:dolphins}, using two different layouts. 
%
\begin{figure}[h!tbp]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/dolphins.pdf}
        \caption{Fruchterman Reingold.}
        %\label{fig:cartAreas1}
    \end{subfigure}%
    \quad
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/dolphins_circle.pdf}
        \caption{Circular.}
        %\label{fig:cartTree1}
    \end{subfigure}
    %(or a blank line to force the subfigure onto a new line)
    \vspace{1\baselineskip}
    \caption{Different layouts for the Dolphins network.}
    \label{fig:dolphins}
\end{figure}

\section{Descriptive Analysis of Network Graph Characteristics}
By descriptive analysis of networks, we usually mean summarizing statistics that can be calculated from a network. These statistics are meant to give insight to the structure of the network. The topic of structural analysis was developed primarily outside 'mainstream' statistics, and as a result, the tools are rather descriptive than inferential.

For large networks some of these statistics are not possible to calculate, and instead they need to be estimated, either through sampling, or through other statistics. However, as our dataset is quite small no estimation was needed.

In this section I used the \verb+igraph+ packages for representing the network in R and calculating the different statistics. As the names of the functions are typically the same as the names of the characteristics, I will not further comment on the individual functions.

\subsection{Vertex Characteristics}
\label{Vertex Characteristics}
When characterizing a network it is common to investigate the degree distribution.
For an undirected graph $G = (V, E)$, the degree of each vertex $v \in V$ is the number of edges connected to $v$. In other words, the degree is the number of neighbors of $v$. The degree distribution is the degree frequencies of the graph, often normalized.

In Figure~\ref{fig:dolph_degree} we have displayed the degree frequencies of the Dolphins network. From the figure, we can see that some dolphins are much more connected than others. This means that it is not one large group where everyone interacts with each other, but instead some dolphins are much more social than others.
\\
\begin{figure}[h!tbp]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/dolph_degree.pdf}
        \caption{Frequencies.}
        \label{fig:dolph_degree}
    \end{subfigure}%
    \quad
    \begin{subfigure}[b]{0.48\textwidth}
        \includegraphics[width=\textwidth]{./figures/dolph_degree_log.pdf}
        \caption{Log scale.}
        \label{fig:dolph_degree_log}
    \end{subfigure}
    %(or a blank line to force the subfigure onto a new line)
    \vspace{1\baselineskip}
    \caption{Degree distribution for the Dolphins network.}
    \label{fig:dolph_degree_all}
\end{figure}
%
\\
In real world networks we often find that the degree distribution follows a power law. This is so common that there the term 'scale-free networks' has been coined to describe such networks. If a network is scale-free, most vertices will have a small degree distribution, while a couple 'hubs' will be very highly connected. It quite common that social networks are scale-free.

In Figure~\ref{fig:dolph_degree_log} we see the log frequencies against the log degrees. The points do not follow a straight line, so there is no indication of a power law here.
\\
\\
Another way to investigate the degrees are to plot each node's degree against the average degree its neighbors. This can possibly reveal patterns in the connectivity. 
Figure~\ref{fig:dolph_degree_neigh} shows this plot for the Dolphins network. 
It might look like highly connected dolphins as a tendency to interact with other highly connected dolphins, while low degree dolphins has a tendency to interact with dolphins of both higher and lower degree.
%
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[scale=0.5]{./figures/dolph_degree_neigh.pdf}
    \end{center}
    \caption{Dolphin dataset. Degrees plotted against the average degree of the vertex's neighbors.}
    \label{fig:dolph_degree_neigh}
\end{figure}

\subsection{Network Cohesion}
Cohesion is concerned with how subsets are cohesive, or 'stuck together'. As an example, one could investigate if friends of friends typically also are friends.

In the Dolphins network, we find that there are 62 nodes,  159 edges, 95 triangles, 27 cliques of size 4, and  3 cliques of size 5. So the largest groups of dolphins that all interact are of size 5.

The density of a graph is defined as number of edges divided by potential number of edges (so a clique has density 1, while no edges refers to density 0). The Dolphins network has density $0.084$, so less than 10 \% of the vertices has an edge between them.

The clustering coefficient, or transitivity, is the ratio of triplets (3 cliques) to the number of 2 stars (three nodes connected by two edges). Thus this gives the proportion of friends friends that are also friends.
For the Dolphins we get a clustering coefficient of $0.31$.

A similar way to investigate cohesiveness is through the shortest paths between vertices. As the transitivity is only concerned with neighbors, the average shortest path gives a different perspective on how closely connected the network is. 
For our dolphins, the average shortest path is 3.36, and the longest shortest path between two dolphins (diameter of network) is 8. Networks with high transitivity, and low diameter are said to have the 'small-world' property. This property is quite common in social networks, but from the analysis so far, it is not immediately clear if the Dolphins network has this property. We will get back to this later on.

\subsubsection{Cuts}
Network cohesion can alternatively be characterized through how fragile the network is to removal of nodes or edges.
For the Dolphins network we find that the removal of a single well-chosen vertex or edge will split the network into multiple components.
There are actually seven dolphins that, when removed, will result in splitting the network. Their average degree is 8.29, so these are some of the more well-connected dolphins.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mathematical Models}

The characteristics we have listed so far sometimes need more context to be informative. We might need to compare their values to other networks to see if they are high or low. One way to do this, is through finding an appropriate network model for describing the network and compare its characteristics to that of the network.
We can typically divide the network models into two set of models; mathematical and statistical. Mathematical network models are different from statistical network models in the sense that they are simpler in nature and more amendable for mathematical analysis. On the other hand, statistical models can often be a better fit of the network, though more complex.

\subsection{Classical and Generalized Random Graph Models}

Previously, we found that the largest clique in the Dolphins network was of size 5, and the network had 3 such cliques. 
These statistics can be compared with a network model by drawing random graphs from the model and compute the same statistics.
We started by simulating 1000 graphs with the same number of vertices and edges as in the Dolphins network. 
All edges were randomly assigned with uniform probability. These graphs are referred to as classical random graphs, and are available in R through the function \verb+erdos.renyi.game+ in the \verb+igraph+ package.

In the simulations we only got graphs with largest clique of 3 or 4. 0.838 of the graphs with clique value 3 and 0.162 with clique value 4. As the classical random graphs do net seem to be a particularly good fit, this indicates that the dolphins associations with each other is not completely random.

By repeating the experiment, but with generalized random graphs that follows the same degree distribution as the dolphins, we get $0.004$ graphs with clique number 5 (3: 0.504, 4:  0.492, 5: 0.004). This means that the dolphins interaction patterns are not fully explained by their degree distribution either.
The generalized random graph was fitted through the \verb+degree.sequence.game+ function in \verb+igraph+.
\\
\\
Using the same simulations as above we can evaluate the clustering coefficient and average shortest path in the dolphins network. The maximum average shortest path in the classical and generalized random graph models were  2.80 and 2.91, respectively. This is quite far from the average shortest path of the dolphins, which was 3.36.  We see the same trend for the maximum clustering coefficients; 0.131 for random graphs, 0.163 for generalized random graph, and 0.309 for the dolphins.
So clearly these models are not able to explain much of the structure in the network at all. 
\\
\\
We previously mentioned that many social networks have the so called small-world property. A common way to assess this property is to compare the clustering coefficient and average shortest path length to that of a classical random graph. If the clustering coefficient exceeds that of a random graph, while the average shortest path length stays roughly the same, the network is said to have the small-world property.

For our Dolphins network, both the clustering coefficient and the average shortest path are both much higher than that of the classical random graph. We therefore conclude that the net work does not have the small-world property.

\subsection{Preferential Attachment Models}
Preferential attachment models represent graphs that are constructed under the assumption that 'the rich get richer'. These models assume that you start with a graph, and when a new vertex is introduced, it is probabilistically connected to existing vertices in coherence with their degree. We will assume the probability of forming an edge is \textit{proportional} to the nodes degree.
It has been found that the preferential attachment models are good at creating power law degree distributions. While we found, in Section~\ref{Vertex Characteristics}, that the dolphins degree distribution does not follow a power law, preferential attachment model might still produce some useful results.
\\
\\
When we simulate a preferential attachment model, we grow the network by adding a new node with $m$ edges, and the edges connect to the existing nodes. Thus, we need to specify $m$ in the model. In the Dolphins network the number of edges relative to vertices is $2.56$, so we choose $m$ as 2 and 3. From the degree distribution in Figure~\ref{fig:dolph_degree}, we see that many of the dolphins only have one edge, suggesting that this model is not a particularly good idea.
We used the \verb+barabasi.game+ function to simulate preferential attachment models.

The results of the simulations can be found in Table~\ref{tab:PAM}. By comparing the results to the dolphins network with largest clique of 5, average path of 3.36, and transitivity of 0.31, we confirm that the preferential attachment models are not particularly good fits.
\\
%
\begin{table}[h!tbp]
    \centering
    \caption{Results from 1000 simulated preferential attachment models.}
    \label{tab:PAM}
    \begin{tabular}{rrrr}
        \hline
         $m$ &  largest clique &  max avg path &  max transitivity  \\
        \hline
        3 &  4 & 2.496 & 0.192 \\
        4 & 4 &  2.498 & 0.186 \\
        \hline
    \end{tabular}
\end{table}
%
\\
We have found no mathematical models that were a good fit for our Dolphins network.
This suggest that there are some other non-trivial process determining the association between the dolphins.

\section{Statistical Models}
While the mathematical models in the previous section try to model different phenomena, statistical models, on the other hand, are created to be fitted to the networks. As a result, the statistical models might not be as intuitive as the mathematical models.

It is common to include exogenous variables (vertex attributes) as covariates for the statistical models. This makes it possible to study these variables' effect on the formation of the network, which can  often be our main interest.
For the Dolphins network it could for instance be interesting to see if gender, age, and size of the dolphins affected their interaction patterns. 
Unfortunately, the dataset contains no exogenous variables, restricting us to only rely on the edge structures in the network for constructing covariates.

\subsection{Exponential Random Graph Models}

Exponential Random Graph Models (ERGM) are designed in direct analogy to the classical generalized linear models (GLM). 
An ERGM is of the form
\begin{align}
    \text P(\mathbf{Y} = \mathbf{y}) = \frac{1}{\kappa} \exp \left\{\sum_H \theta_H g_H(\mathbf{y})\right\},
\end{align}
where $\mathbf{Y}$ denotes a (random) adjacency matrix for the graph $G$, $H$ is a configuration of edges, $g_H(\mathbf y)$ is an indicator of $H$ occurring in  $\mathbf y$, and $\theta_H$ is a parameter for the configuration $H$.

When constructing an ERGM, we find a set of configurations $H$, e.g. edges, triangles and k-stars, and fit the $\theta_H$ parameter by maximizing the log-likelihood.
Having said that, this often produce models that put a disproportionate weight on a small set of outcomes, resulting in a bad fit. As a result, other statistics that attempt to summarize this information has been proposed. Alternating k-star statistic, geometrically weighted degree count, and geometrically weighted edgewise shared partner distribution are three such statistics. 
I was only able to fit the geometrically weighted degree count to the Dolphins network, as the other two statistics reported an error during the MCMC iterations.
However, as long as the edges are included in the model, alternating k-stars is equivalent to the geometrically weighted degree count (though their interpretation is not).

The geometrically weighted degree count, or GWD, is defined as
\begin{align}
    \text{GWD}_\gamma (\mathbf y) = \sum_{d = 0}^{N_v - 1} e^{-\gamma d}N_d(\mathbf y),
\end{align}
where $N_d(\mathbf y)$ is the number of vertices with degree $d$. In a sense, this approach attempts to model the degree distribution, where $\gamma > 0$ influences the extent to which higher degree nodes are likely to occur in the graph.

An ERGM was fitted to the network, using edges and GWD, and  we let the \verb+ergm+ package estimate the $\gamma$ parameter. 
The output under summarize the results.
I do not think the \verb+dwdegree.decay+ is the same parameter as $\gamma$, though I do think they are somewhat related.  
This was hard to understand from the documentation of \verb+ergm+. 
\begin{verbatim}
    ==========================
    Summary of model fit
    ==========================

    Formula:   g.s ~ edges + gwdegree(1, fixed = FALSE)

    Iterations:  2 out of 20

    Monte Carlo MLE Results:
                   Estimate Std. Error MCMC %  p-value    
    edges           -1.8209     0.2580      0  < 1e-04 ***
    gwdegree        -1.6541     0.4789      0 0.000564 ***
    gwdegree.decay   0.9971     0.5929      0 0.092774 .  
    ---
    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

    Null Deviance: 2621  on 1891  degrees of freedom
    Residual Deviance: 1085  on 1888  degrees of freedom

    AIC: 1091    BIC: 1107    (Smaller is better.) 
\end{verbatim}
Here we see that both the edges and the GWD are considered significant, though it is important to note that the theoretical justification for the asymptotic chi-squared and $F$-distributions used by \verb+ergm+ has not yet been established.

The estimated coefficients can be interpreted as the log-odds of an edge being present, conditioned on the rest of the network. As the GWD and edge count are not vertex attributes, interpretation of their coefficients is hard. Therefore I will only say that the significance of GWD indicate that the degree distribution is important when modeling the dolphins interaction pattern.

I have not found any source that gives me a good explanation of the \verb+gwdegree.decay+ parameter, so I will not try to interpret it.
\\
\\
When we worked with the mathematical models, we could easily draw from the specified models and compare characteristics with the Dolphins network.
ERGMs on the other hand, are probabilistic models, and not specified though a growth model. This makes them harder to simulate. There are, however, some functionality in the \verb+ergm+ library that can help us with this job. 
When we continue to work with other statistical models, we will see that this simulation becomes harder, mainly because of lack of implemented methods in R. Therefore we will not be able to calculate the same types of characteristics as for the mathematical models, and I have instead chosen to focus on the packages' implementations for evaluating goodness of fit.

\subsubsection{Goodness of Fit}
To measure how well our fit is, we use the supplied \verb+gof.ergm+ function that simulates from the fitted model and compares different characteristics with the original network. This function is not very well documented, but it is suppose to run the ''necessary Monte Carlo simulation and calculates comparisons with the original network graph in terms of the distribution of degree, geodesic length (shortest path between nodes), and edge-wise shared partners (i.e., the number of neighbors shared by a pair of vertices defining an edge)''\citep{dolphins}.
The plots can be found in Figure~\ref{fig:gof_ergm}.
The ERGM model seems to have captured the degree distribution quite well. On the other hand, the geodesic length and the edge-wise shared partners are not very similar to the Dolphins network.
%
\begin{figure}[h!tbp]
    \begin{center}
        \includegraphics[scale=0.9]{./figures/gof_ergm.pdf}
    \end{center}
    \caption{Goodness of fit plots for ERGM model fitted to the Dolphins network. The three figures display the degree distribution, the edge-wise shared partners, and geodesic distance. The bold solid lines represent the original network, while the box-plots and the 10th and 90th quantile curves are from the MC simulations.}
    \label{fig:gof_ergm}
\end{figure}

\subsection{Network Block Models}
While ERGMs were analogous to GLM, network block models are analogous to classical mixture models. Standard block models assume that we have $Q$ classes, and we know which of our nodes are in each class. In our setting, we have no such classes, so instead we investigate stochastic block models, which only assumes knowledge of the total number of classes.

The regular block models can be expressed as an ERGM through
\begin{align}
    \label{eq:block_model}
    \text P(\mathbf{Y} = \mathbf{y}) = \frac{1}{\kappa} \exp \left\{\sum_{q, r} \theta_{q, r} L_{q, r}(\mathbf{y})\right\},
\end{align}
where $L_{q, r}(\mathbf{y})$ is the number of edges in the observed graph $\mathbf y$ connecting pairs of vertices of class $q$ and $r$. 
Thus, the bock model is just a variant of the Bernoulli random graph model, with $Q^2$ edge probabilities $\pi_{q, r}$.

The stochastic version assumes in addition that the nodes are randomly assigned to the classes. This is specified through the indicator variables $Z_{iq}$ relating node $i$ to class $q$. The $Z$'s are determined independently with $\text P(Z_{iq}) = \alpha_q$, and conditioned on the $Z$'s the edges are modeled though \eqref{eq:block_model}.
Thus, a stochastic block model is a mixture of ERGMs.

We fitted a stochastic block models to our Dolphins data, using the \verb+mixer+ library in R. It suggested to divide the dolphins into three classes, with estimated proportions $\alpha_q$ of 0.482, 0.353, and  0.165. So one class should contain almost half of the dolphins.

\subsubsection{Goodness of Fit}
I did not find an implementation of how to simulate from the fitted model. 
As I was able to sample from an ERGM previously, I could possibly draw from a mixture distribution, and construct the corresponding ERGM through the \verb+gof.ergm+. However, it would require some work, and I have not quite understood how \verb+gof.ergm+ works, so it might not yield the results I think it will.
In addition, in the next section, we work with a model I was not able to sample from, so I had to do alternative analysis there as well.

The \verb+mixture+ packages supplies a set of plots displayed in Figure~\ref{fig:block_models}. At the top left, one can find the integrated classification likelihood, which is a measure similar to AIC and BIC, but adapted for clustering problems. This criterion choose 3 classes as the optimum. On the top right of Figure~\ref{fig:block_models}, we can find a reorganized adjacency matrix. This shows to what extent edges are within classes and across classes. Much of the same is displayed in the bottom right plot, where the width of the edges corresponds to the number of edges connecting the classes. From both plots we see that the network is partitioned such that two of the classes have no edges between them. 
Finally, on the bottom left, we find the degree distribution, where the yellow histogram are the dolphins network, and the blue line is the fitted model. It does not seem like the degree distribution is particularly accurate.
%
\begin{figure}[h!tbp]
    \begin{center}
        \includegraphics[scale=0.9]{./figures/block_models.pdf}
    \end{center}
    \caption{Various plots summarizing the goodness of fit for the stochastic block model analysis of the Dolphins network}
    \label{fig:block_models}
\end{figure}

\subsection{Latent Network Models}
Latent network models are models that incorporate latent variables as part of the model specification. The stochastic block models utilize the unknown class affiliation as latent variables. Now, we will consider latent network models of the form 
\begin{align}
    \label{eq:latent}
    \text P(Y_{ij} = 1 \mid \mathbf{X}_{ij} = \mathbf{x}_{ij}) = \Phi (\mu + \mathbf x_{ij}^T \mathbf \beta + \alpha(u_i, u_j)),
\end{align}
where $\Phi$ is the cumulative distribution function of a standard normal distribution, $\mu$ is a constant, the $u_i$'s are independent and identically distributed latent variables, $\mathbf x_{ij}$ are pair specific covariates, and $\alpha$ is a symmetric function.
However, we do not have any pair specific covariates, so we our model is just a function of the latent variables and $\mu$.

$\alpha$ can be specified in different ways, but here we only use the form $\alpha(u_i, u_j) = u_i^T \Lambda u_j$, where $\Lambda$ is a $Q \times Q$ matrix. This is called the eigenmodel.
\\
\\
We fitted an eigenmodel to our Dolphins network, using the library \verb+eigenmodel+, and chose $Q=3$ as we found three mixtures in the previous section.

It should be easy so sample from the model. We just need $\mu$ and $\alpha(u_i, u_j)$ to specify the probabilities in \eqref{eq:latent}.  
As the edge assignment in this latent model is independent of the other edges, we could just sample edges from a Bernoulli distribution with probabilities specified by \eqref{eq:latent}. However, I was not able to obtain $\mu$.
So instead, to evaluate the fit, I followed the analysis in the book \cite{kolaczyk2014}. Here we fit a model to a subset of the network, and make the model predict the rest of the network. This was done using 5-fold cross-validation.

A common evaluation metric is the ROC AUC, which is the area under the ROC curve of the predictions. Random assignment will results in AUC of approximately 0.5, while perfect classification will give AUC of 1. From our experiment we got a AUC of 0.54, so the model performs only slightly better than random edge assignment.
This does not necessarily mean the model is a bad fit, it might just be that the network is hard to predict. For instance, edge prediction of a classical random graph would result is the same probability of all edges, and an AUC of $0.5$.

\section{Summary}
In this project we have looked at the social network of dolphins by \cite{dolphins}. We found that it was neither scale-free, nor had the small-world property. We could also not find any model that that was able to describe the social structure particularly well.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cleardoublepage{}
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
%\bibliographystyle{plain}
%\bibliographystyle{apa}
%\bibliographystyle{authordate4}
%\bibliographystyle{ksfh_nat}
\bibliographystyle{plainnat}
\bibliography{bibliography}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\clearpage
%\begin{appendices}
  %% Use \section
  %% Can use \phantomsection in text to get back
%\end{appendices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{thebibliography}{99} % At most 99 references.
%% Use "bibit" to generate bibitem
%\end{thebibliography}
%
\end{document}

### 2016-11-23 latex/compile.sh

@@ -1,0 +1,5 @@
+#!/usr/bin/env bash
+
+bibtex article.aux
+
+pdflatex article.tex

### 2016-11-23 project_code.R

@@ -1,3 +1,13 @@
+
+# Printing pdf figures
+printfig <- function(name, height = 6, ...){
+  path <- paste('~/stk4080//latex/figures/', name, '.pdf', sep = '')
+  pdf(file = path, height = height, ...)
+  par(mai = c(0.6, 0.6, 0.1, 0.1)) # manipulate inner margins of subplots
+  par(mgp = c(1.8, 0.7, 0)) # manipulate spacing of axis ticks, labels, text
+  par(omi = c(0, 0, 0, 0)) # manipulate outer margin of full plot
+}
+
 pbc <- read.csv("~/stk4080/pbc3A.txt", sep="")
 head(pbc)
 # sapply(pbc, class)
@@ -8,11 +18,12 @@ head(pbc)
 # plot(pbc$beh, pbc$dod)
 # plot(pbc)
 # plot(pbc$kjonn, pbc$ald)
-# 
+ 
+
 # Not unique death / censor times
+# We have tied data because of rounding to nearest day.
 length(pbc$dager) 
 length(unique(pbc$dager))
-
 #----------
 # a) Enkle univariate analyser:
 # Først utfører du enkle univariate analyser ved hjelp av Kaplan-Meier plot og 
@@ -24,61 +35,190 @@ length(unique(pbc$dager))
 # Kaplan-Meier
 #---------------
 library(survival)
+printfig('km_beh', height = 5, width=5)
+layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
 fit = survfit(Surv(dager, dod)~beh, data = pbc)
+survfit(Surv(pbc$dager, pbc$dod)~pbc$beh)
 plot(fit, lty=1:2, ylab='survival', xlab='days')
 legend(1,0.2,c("treatment","placebo"), lty=1:2, bty='n')
 grid()
+plot(as.factor(pbc$beh), main='', ylab='frequency', xlab='treatment', col='white',
+     names.arg=c('treatment', 'placebo'))
+dev.off()
+
+survdiff(Surv(dager, dod)~beh, data = pbc)
 
 library(dplyr)
 df = tbl_df(pbc)
 glimpse(df)
 df %>% group_by(kjonn) %>% summarise(avg = mean(dager))
 
+str_lab = function(gr){
+  gr = as.character(gr)
+  lab = c(paste('-', gr[2]))
+  for (i in 2:(length(gr)-1)){
[…]

### 2016-11-25 latex/article.tex
More work

@@
     \end{center}
-    \vspace{-0.8cm}
+    \vspace{-0.2cm}
     \caption{Martingale residuals for cox regression on age.}
@@
 \subsection{Bilirubin}
+
+We fit a cox regression to the bilirubin covariate, and the results can be found in Table~\ref{tab:cox_bilirubin}, and the Martingale residuals in Figure~\ref{fig:cox_bil_mart}.
+The score test give a $p$-value of 0, but the we see a clear pattern in the martingale residuals.
+We therefore fit a cox regression with a cubic smoothing-spline to the bilirubin covariate, and got that both the linear and non-linear part was highly significant ($p < 1e-9$, table exempt).
+
+The shape of the martingale residuals might suggest that a log-transform of the bilirubin variable could be a better fit.
+We also plot the spline term against the amount of bilirubin in Figure~\ref{fig:cox_bil_termplot}, where the x-axis is logarithmically scaled.
+We see that the line is almost linear, supporting our claim that a log-transform might be suitable.
+
+Next we fitted a cox regression to the log transformed bilirubin covariates, yielding the results in Table~\ref{tab:cox_bilirubin_log}.
+
+The martingale residuals were plotted in Figure~\ref{fig:cox_bil_mart_log}, where the residuals are evenly distributed around zero. This again support our log-transform.
+
 \input{tables/cox_bilirubin}
+\unsure[inline]{Something on how log transform is reasonable.}
+
+\begin{figure}[h!tb]
+    \begin{center}
+        \includegraphics[scale=0.6]{./figures/cox_bil_mart.pdf}
+    \end{center}
+    \vspace{-0.2cm}
+    \caption{Martingale residuals for cox regression on bilirubin.}
+    \label{fig:cox_bil_mart}
+\end{figure}
+
+\begin{figure}[h!tbp]
+    \centering
+    \begin{subfigure}[b]{0.48\textwidth}
+        \includegraphics[width=\textwidth]{./figures/cox_bil_termplot.pdf}
+        \caption{Plot of cox regression term for cubic smoothing-spline. Not that the x-axis is logarithmically scaled.}
+        \label{fig:cox_bil_termplot}
+    \end{subfigure}%
+    \quad
+    \begin{subfigure}[b]{0.48\textwidth}
+        \includegraphics[width=\textwidth]{./figures/cox_bil_mart_log.pdf}
+        \caption{Martingale residuals of cox regression on log bilirubin, and GAM fitted to the residuals.}
+        \label{fig:cox_bil_mart_log}
+    \end{subfigure}
+    %(or a blank line to force the subfigure onto a new line)
+    \vspace{1\baselineskip}
+    \caption{}
+    \label{fig:cox_bil_term_and_mart}
+\end{figure}
+
+\input{tables/cox_bilirubin_log}
 
 \subsection{Albumin}
+\input{tables/cox_albumin}
+
+\begin{figure}[h!tb]
+    \begin{center}
+        \includegraphics[scale=0.6]{./figures/cox_alb_mart.pdf}
+    \end{center}
+    \vspace{-0.2cm}
+    \caption{Martingale residuals for cox regression on albumin.}
+    \label{fig:cox_alb_mart}
+\end{figure}
 
@@
 
-\clearpage
-\section{Dolphin Stuff}
-
-\subsection{The Dataset}
-
-All analysis was done on a directed social network of bottlenose dolphins \citep{dolphins}.
-The nodes are the bottlenose dolphins (genus Tursiops) of a bottlenose dolphin community living off Doubtful Sound, a fjord in New Zealand (spelled fiord in New Zealand).
-An edge indicates a frequent association.
-The dolphins were observed between 1994 and 2001.
-
-Part of what could make this network interesting, is that we can study the social interactions between highly intelligent animal other than humans.
-So we could for instance compare the dolphins to different species, to investigate the impact of intelligence on social structures.
-However, I will do no such analysis, and I will admit that the network was chosen partly because of its size.
-It only contains 62 dolphins, so much of the analysis is not particularly computationally expensive.
-The network is displayed in Figure~\ref{fig:dolphins}, using two different layouts. 
+We see from Figure~\ref{fig:cor_heat} that the covariates are not particularly correlated.
+The highest (in absolute value) is between bilirubin and albumin on approximately $-0.4$.
+And from a quick look at all the covariates plotted against each other, there does not seem to be any non-linear effect that cause the low correlations (figures not included).
 %
-\begin{figure}[h!tbp]
-    \centering
-    \begin{subfigure}[b]{0.48\textwidth}
-        \includegraphics[width=\textwidth]{./figures/dolphins.pdf}
-        \caption{Fruchterman Reingold.}
-        %\label{fig:cartAreas1}
-    \end{subfigure}%
-    \quad
-    \begin{subfigure}[b]{0.48\textwidth}
-        \includegraphics[width=\textwidth]{./figures/dolphins_circle.pdf}
-        \caption{Circular.}
-        %\label{fig:cartTree1}
-    \end{subfigure}
-    %(or a blank line to force the subfigure onto a new line)
-    \vspace{1\baselineskip}
-    \caption{Different layouts for the Dolphins network.}
-    \label{fig:dolphins}
+\begin{figure}[h!tb]
+    \begin{center}
+        \includegraphics[scale=0.6]{./figures/cor_heat.pdf}
+    \end{center}
+    \vspace{-0.8cm}
+    \caption{Correlations.}
+    \label{fig:cor_heat}
 \end{figure}
 
-\section{Descriptive Analysis of Network Graph Characteristics}
-By descriptive analysis of networks, we usually mean summarizing statistics that can be calculated from a network.
-These statistics are meant to give insight to the structure of the network.
-The topic of structural analysis was developed primarily outside 'mainstream' statistics, and as a result, the tools are rather descriptive than inferential.
+We started with all covariates and removed the ones that were not significant (just treatment).
 
-For large networks some of these statistics are not possible to calculate, and instead they need to be estimated, either through sampling, or through other statistics.
-However, as our dataset is quite small no estimation was needed.
-
-In this section I used the \verb+igraph+ packages for representing the network in R and calculating the different statistics.
-As the names of the functions are typically the same as the names of the characteristics, I will not further comment on the individual functions.
-
-\subsection{Vertex Characteristics}
-\label{Vertex Characteristics}
-When characterizing a network it is common to investigate the degree distribution.
-For an undirected graph $G = (V, E)$, the degree of each vertex $v \in V$ is the number of edges connected to $v$.
-In other words, the degree is the number of neighbors of $v$.
-The degree distribution is the degree frequencies of the graph, often normalized.
-
-In Figure~\ref{fig:dolph_degree} we have displayed the degree frequencies of the Dolphins network.
-From the figure, we can see that some dolphins are much more connected than others.
-This means that it is not one large group where everyone interacts with each other, but instead some dolphins are much more social than others.
-\\
-\begin{figure}[h!tbp]
-    \centering
-    \begin{subfigure}[b]{0.48\textwidth}
-        \includegraphics[width=\textwidth]{./figures/dolph_degree.pdf}
-        \caption{Frequencies.}
-        \label{fig:dolph_degree}
-    \end{subfigure}%
-    \quad
-    \begin{subfigure}[b]{0.48\textwidth}
-        \includegraphics[width=\textwidth]{./figures/dolph_degree_log.pdf}
-        \caption{Log scale.}
-        \label{fig:dolph_degree_log}
-    \end{subfigure}
-    %(or a blank line to force the subfigure onto a new line)
-    \vspace{1\baselineskip}
-    \caption{Degree distribution for the Dolphins network.}
-    \label{fig:dolph_degree_all}
-\end{figure}
-%
-\\
-In real world networks we often find that the degree distribution follows a power law.
-This is so common that there the term 'scale-free networks' has been coined to describe such networks.
-If a network is scale-free, most vertices will have a small degree distribution, while a couple 'hubs' will be very highly connected.
-It quite common that social networks are scale-free.
-
-In Figure~\ref{fig:dolph_degree_log} we see the log frequencies against the log degrees. The points do not follow a straight line, so there is no indication of a power law here.
-\\
-\\
-Another way to investigate the degrees are to plot each node's degree against the average degree its neighbors. This can possibly reveal patterns in the connectivity. 
-Figure~\ref{fig:dolph_degree_neigh} shows this plot for the Dolphins network. 
-It might look like highly connected dolphins as a tendency to interact with other highly connected dolphins, while low degree dolphins has a tendency to interact with dolphins of both higher and lower degree.
-%
-\begin{figure}[htbp]
-    \begin{center}
-        \includegraphics[scale=0.5]{./figures/dolph_degree_neigh.pdf}
-    \end{center}
-    \caption{Dolphin dataset. Degrees plotted against the average degree of the vertex's neighbors.}
-    \label{fig:dolph_degree_neigh}
-\end{figure}
-
-\subsection{Network Cohesion}
-Cohesion is concerned with how subsets are cohesive, or 'stuck together'. As an example, one could investigate if friends of friends typically also are friends.
-
-In the Dolphins network, we find that there are 62 nodes,  159 edges, 95 triangles, 27 cliques of size 4, and  3 cliques of size 5. So the largest groups of dolphins that all interact are of size 5.
-
-The density of a graph is defined as number of edges divided by potential number of edges (so a clique has density 1, while no edges refers to density 0).
-The Dolphins network has density $0.084$, so less than 10 \% of the vertices has an edge between them.
-
-The clustering coefficient, or transitivity, is the ratio of triplets (3 cliques) to the number of 2 stars (three nodes connected by two edges).
-Thus this gives the proportion of friends friends that are also friends.
-For the Dolphins we get a clustering coefficient of $0.31$.
-
-A similar way to investigate cohesiveness is through the shortest paths between vertices.
-As the transitivity is only concerned with neighbors, the average shortest path gives a different perspective on how closely connected the network is.
-For our dolphins, the average shortest path is 3.36, and the longest shortest path between two dolphins (diameter of network) is 8.
-Networks with high transitivity, and low diameter are said to have the 'small-world' property.
-This property is quite common in social networks, but from the analysis so far, it is not immediately clear if the Dolphins network has this property.
-We will get back to this later on.
-
-\subsubsection{Cuts}
-Network cohesion can alternatively be characterized through how fragile the network is to removal of nodes or edges.
-For the Dolphins network we find that the removal of a single well-chosen vertex or edge will split the network into multiple components.
-There are actually seven dolphins that, when removed, will result in splitting the network. Their average degree is 8.29, so these are some of the more well-connected dolphins.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Mathematical Models}
-
-The characteristics we have listed so far sometimes need more context to be informative.
-We might need to compare their values to other networks to see if they are high or low.
-One way to do this, is through finding an appropriate network model for describing the network and compare its characteristics to that of the network.
-We can typically divide the network models into two set of models; mathematical and statistical.
-Mathematical network models are different from statistical network models in the sense that they are simpler in nature and more amendable for mathematical analysis.
-On the other hand, statistical models can often be a better fit of the network, though more complex.
-
-\subsection{Classical and Generalized Random Graph Models}
-
-Previously, we found that the largest clique in the Dolphins network was of size 5, and the network had 3 such cliques. 
-These statistics can be compared with a network model by drawing random graphs from the model and compute the same statistics.
-We started by simulating 1000 graphs with the same number of vertices and edges as in the Dolphins network. 
-All edges were randomly assigned with uniform probability.
-These graphs are referred to as classical random graphs, and are available in R through the function \verb+erdos.renyi.game+ in the \verb+igraph+ package.
-
-In the simulations we only got graphs with largest clique of 3 or 4.
-0.838 of the graphs with clique value 3 and 0.162 with clique value 4.
-As the classical random graphs do net seem to be a particularly good fit, this indicates that the dolphins associations with each other is not completely random.
-
-By repeating the experiment, but with generalized random graphs that follows the same degree distribution as the dolphins, we get $0.004$ graphs with clique number 5 (3: 0.504, 4:  0.492, 5: 0.004).
-This means that the dolphins interaction patterns are not fully explained by their degree distribution either.
-The generalized random graph was fitted through the \verb+degree.sequence.game+ function in \verb+igraph+.
-\\
-\\
-Using the same simulations as above we can evaluate the clustering coefficient and average shortest path in the dolphins network.
-The maximum average shortest path in the classical and generalized random graph models were  2.80 and 2.91, respectively.
-This is quite far from the average shortest path of the dolphins, which was 3.36.
-We see the same trend for the maximum clustering coefficients; 0.131 for random graphs, 0.163 for generalized random graph, and 0.309 for the dolphins.
-So clearly these models are not able to explain much of the structure in the network at all. 
-\\
-\\
-We previously mentioned that many social networks have the so called small-world property.
-A common way to assess this property is to compare the clustering coefficient and average shortest path length to that of a classical random graph.
-If the clustering coefficient exceeds that of a random graph, while the average shortest path length stays roughly the same, the network is said to have the small-world property.
-
-For our Dolphins network, both the clustering coefficient and the average shortest path are both much higher than that of the classical random graph.
-We therefore conclude that the net work does not have the small-world property.
-
-\subsection{Preferential Attachment Models}
-Preferential attachment models represent graphs that are constructed under the assumption that 'the rich get richer'.
-These models assume that you start with a graph, and when a new vertex is introduced, it is probabilistically connected to existing vertices in coherence with their degree.
-We will assume the probability of forming an edge is \textit{proportional} to the nodes degree.
-It has been found that the preferential attachment models are good at creating power law degree distributions.
-While we found, in Section~\ref{Vertex Characteristics}, that the dolphins degree distribution does not follow a power law, preferential attachment model might still produce some useful results.
-\\
-\\
-When we simulate a preferential attachment model, we grow the network by adding a new node with $m$ edges, and the edges connect to the existing nodes.
-Thus, we need to specify $m$ in the model.
-In the Dolphins network the number of edges relative to vertices is $2.56$, so we choose $m$ as 2 and 3.
-From the degree distribution in Figure~\ref{fig:dolph_degree}, we see that many of the dolphins only have one edge, suggesting that this model is not a particularly good idea.
-We used the \verb+barabasi.game+ function to simulate preferential attachment models.
-
-The results of the simulations can be found in Table~\ref{tab:PAM}.
-By comparing the results to the dolphins network with largest clique of 5, average path of 3.36, and transitivity of 0.31, we confirm that the preferential attachment models are not particularly good fits.
-\\
-%
-\begin{table}[h!tbp]
-    \centering
-    \caption{Results from 1000 simulated preferential attachment models.}
-    \label{tab:PAM}
-    \begin{tabular}{rrrr}
-        \hline
-         $m$ &  largest clique &  max avg path &  max transitivity  \\
-        \hline
-        3 &  4 & 2.496 & 0.192 \\
-        4 & 4 &  2.498 & 0.186 \\
-        \hline
-    \end{tabular}
-\end{table}
-%
-\\
-We have found no mathematical models that were a good fit for our Dolphins network.
-This suggest that there are some other non-trivial process determining the association between the dolphins.
-
-\section{Statistical Models}
-While the mathematical models in the previous section try to model different phenomena, statistical models, on the other hand, are created to be fitted to the networks.
-As a result, the statistical models might not be as intuitive as the mathematical models.
-
-It is common to include exogenous variables (vertex attributes) as covariates for the statistical models.
-This makes it possible to study these variables' effect on the formation of the network, which can  often be our main interest.
-For the Dolphins network it could for instance be interesting to see if gender, age, and size of the dolphins affected their interaction patterns. 
-Unfortunately, the dataset contains no exogenous variables, restricting us to only rely on the edge structures in the network for constructing covariates.
-
-\subsection{Exponential Random Graph Models}
-
-Exponential Random Graph Models (ERGM) are designed in direct analogy to the classical generalized linear models (GLM). 
-An ERGM is of the form
-\begin{align}
-    \text P(\mathbf{Y} = \mathbf{y}) = \frac{1}{\kappa} \exp \left\{\sum_H \theta_H g_H(\mathbf{y})\right\},
-\end{align}
-where $\mathbf{Y}$ denotes a (random) adjacency matrix for the graph $G$, $H$ is a configuration of edges, $g_H(\mathbf y)$ is an indicator of $H$ occurring in  $\mathbf y$, and $\theta_H$ is a parameter for the configuration $H$.
-
-When constructing an ERGM, we find a set of configurations $H$, e.g. edges, triangles and k-stars, and fit the $\theta_H$ parameter by maximizing the log-likelihood.
-Having said that, this often produce models that put a disproportionate weight on a small set of outcomes, resulting in a bad fit.
-As a result, other statistics that attempt to summarize this information has been proposed.
-Alternating k-star statistic, geometrically weighted degree count, and geometrically weighted edgewise shared partner distribution are three such statistics.
-I was only able to fit the geometrically weighted degree count to the Dolphins network, as the other two statistics reported an error during the MCMC iterations.
-However, as long as the edges are included in the model, alternating k-stars is equivalent to the geometrically weighted degree count (though their interpretation is not).
-
-The geometrically weighted degree count, or GWD, is defined as
-\begin{align}
-    \text{GWD}_\gamma (\mathbf y) = \sum_{d = 0}^{N_v - 1} e^{-\gamma d}N_d(\mathbf y),
-\end{align}
-where $N_d(\mathbf y)$ is the number of vertices with degree $d$.
-In a sense, this approach attempts to model the degree distribution, where $\gamma > 0$ influences the extent to which higher degree nodes are likely to occur in the graph.
-
-An ERGM was fitted to the network, using edges and GWD, and  we let the \verb+ergm+ package estimate the $\gamma$ parameter. 
-The output under summarize the results.
-I do not think the \verb+dwdegree.decay+ is the same parameter as $\gamma$, though I do think they are somewhat related.  
-This was hard to understand from the documentation of \verb+ergm+. 
-\begin{verbatim}
-    ==========================
-    Summary of model fit
-    ==========================
-
-    Formula:   g.s ~ edges + gwdegree(1, fixed = FALSE)
-
-    Iterations:  2 out of 20
-
-    Monte Carlo MLE Results:
-                   Estimate Std. Error MCMC %  p-value    
-    edges           -1.8209     0.2580      0  < 1e-04 ***
-    gwdegree        -1.6541     0.4789      0 0.000564 ***
-    gwdegree.decay   0.9971     0.5929      0 0.092774 .  
-    ---
-    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
-
-    Null Deviance: 2621  on 1891  degrees of freedom
-    Residual Deviance: 1085  on 1888  degrees of freedom
-
-    AIC: 1091    BIC: 1107    (Smaller is better.) 
-\end{verbatim}
-Here we see that both the edges and the GWD are considered significant, though it is important to note that the theoretical justification for the asymptotic chi-squared and $F$-distributions used by \verb+ergm+ has not yet been established.
-
-The estimated coefficients can be interpreted as the log-odds of an edge being present, conditioned on the rest of the network.
-As the GWD and edge count are not vertex attributes, interpretation of their coefficients is hard.
-Therefore I will only say that the significance of GWD indicate that the degree distribution is important when modeling the dolphins interaction pattern.
-
-I have not found any source that gives me a good explanation of the \verb+gwdegree.decay+ parameter, so I will not try to interpret it.
-\\
-\\
-When we worked with the mathematical models, we could easily draw from the specified models and compare characteristics with the Dolphins network.
-ERGMs on the other hand, are probabilistic models, and not specified though a growth model.
-This makes them harder to simulate.
-There are, however, some functionality in the \verb+ergm+ library that can help us with this job.
-When we continue to work with other statistical models, we will see that this simulation becomes harder, mainly because of lack of implemented methods in R.
-Therefore we will not be able to calculate the same types of characteristics as for the mathematical models, and I have instead chosen to focus on the packages' implementations for evaluating goodness of fit.
-
-\subsubsection{Goodness of Fit}
-To measure how well our fit is, we use the supplied \verb+gof.ergm+ function that simulates from the fitted model and compares different characteristics with the original network.
-This function is not very well documented, but it is suppose to run the ''necessary Monte Carlo simulation and calculates comparisons with the original network graph in terms of the distribution of degree, geodesic length (shortest path between nodes), and edge-wise shared partners (i.e., the number of neighbors shared by a pair of vertices defining an edge)''\citep{dolphins}.
-The plots can be found in Figure~\ref{fig:gof_ergm}.
-The ERGM model seems to have captured the degree distribution quite well. On the other hand, the geodesic length and the edge-wise shared partners are not very similar to the Dolphins network.
-%
-\begin{figure}[h!tbp]
-    \begin{center}
-        \includegraphics[scale=0.9]{./figures/gof_ergm.pdf}
-    \end{center}
-    \caption{Goodness of fit plots for ERGM model fitted to the Dolphins network.
-The three figures display the degree distribution, the edge-wise shared partners, and geodesic distance.
-The bold solid lines represent the original network, while the box-plots and the 10th and 90th quantile curves are from the MC simulations.}
-    \label{fig:gof_ergm}
-\end{figure}
-
-\subsection{Network Block Models}
-While ERGMs were analogous to GLM, network block models are analogous to classical mixture models.
-Standard block models assume that we have $Q$ classes, and we know which of our nodes are in each class.
-In our setting, we have no such classes, so instead we investigate stochastic block models, which only assumes knowledge of the total number of classes.
-
-The regular block models can be expressed as an ERGM through
-\begin{align}
-    \label{eq:block_model}
-    \text P(\mathbf{Y} = \mathbf{y}) = \frac{1}{\kappa} \exp \left\{\sum_{q, r} \theta_{q, r} L_{q, r}(\mathbf{y})\right\},
-\end{align}
-where $L_{q, r}(\mathbf{y})$ is the number of edges in the observed graph $\mathbf y$ connecting pairs of vertices of class $q$ and $r$. 
-Thus, the bock model is just a variant of the Bernoulli random graph model, with $Q^2$ edge probabilities $\pi_{q, r}$.
-
-The stochastic version assumes in addition that the nodes are randomly assigned to the classes.
-This is specified through the indicator variables $Z_{iq}$ relating node $i$ to class $q$.
-The $Z$'s are determined independently with $\text P(Z_{iq}) = \alpha_q$, and conditioned on the $Z$'s the edges are modeled though \eqref{eq:block_model}.
-Thus, a stochastic block model is a mixture of ERGMs.
-
-We fitted a stochastic block models to our Dolphins data, using the \verb+mixer+ library in R.
-It suggested to divide the dolphins into three classes, with estimated proportions $\alpha_q$ of 0.482, 0.353, and  0.165.
-So one class should contain almost half of the dolphins.
-
-\subsubsection{Goodness of Fit}
-I did not find an implementation of how to simulate from the fitted model. 
-As I was able to sample from an ERGM previously, I could possibly draw from a mixture distribution, and construct the corresponding ERGM through the \verb+gof.ergm+.
-However, it would require some work, and I have not quite understood how \verb+gof.ergm+ works, so it might not yield the results I think it will.
-In addition, in the next section, we work with a model I was not able to sample from, so I had to do alternative analysis there as well.
-
-The \verb+mixture+ packages supplies a set of plots displayed in Figure~\ref{fig:block_models}.
-At the top left, one can find the integrated classification likelihood, which is a measure similar to AIC and BIC, but adapted for clustering problems.
-This criterion choose 3 classes as the optimum.
-On the top right of Figure~\ref{fig:block_models}, we can find a reorganized adjacency matrix.
-This shows to what extent edges are within classes and across classes.
-Much of the same is displayed in the bottom right plot, where the width of the edges corresponds to the number of edges connecting the classes.
-From both plots we see that the network is partitioned such that two of the classes have no edges between them.
-Finally, on the bottom left, we find the degree distribution, where the yellow histogram are the dolphins network, and the blue line is the fitted model.
-It does not seem like the degree distribution is particularly accurate.
-%
-\begin{figure}[h!tbp]
-    \begin{center}
-        \includegraphics[scale=0.9]{./figures/block_models.pdf}
-    \end{center}
-    \caption{Various plots summarizing the goodness of fit for the stochastic block model analysis of the Dolphins network}
-    \label{fig:block_models}
-\end{figure}
-
-\subsection{Latent Network Models}
-Latent network models are models that incorporate latent variables as part of the model specification.
-The stochastic block models utilize the unknown class affiliation as latent variables.
-Now, we will consider latent network models of the form
-\begin{align}
-    \label{eq:latent}
-    \text P(Y_{ij} = 1 \mid \mathbf{X}_{ij} = \mathbf{x}_{ij}) = \Phi (\mu + \mathbf x_{ij}^T \mathbf \beta + \alpha(u_i, u_j)),
-\end{align}
-where $\Phi$ is the cumulative distribution function of a standard normal distribution, $\mu$ is a constant, the $u_i$'s are independent and identically distributed latent variables, $\mathbf x_{ij}$ are pair specific covariates, and $\alpha$ is a symmetric function.
-However, we do not have any pair specific covariates, so we our model is just a function of the latent variables and $\mu$.
-
-$\alpha$ can be specified in different ways, but here we only use the form $\alpha(u_i, u_j) = u_i^T \Lambda u_j$, where $\Lambda$ is a $Q \times Q$ matrix. This is called the eigenmodel.
-\\
-\\
-We fitted an eigenmodel to our Dolphins network, using the library \verb+eigenmodel+, and chose $Q=3$ as we found three mixtures in the previous section.
-
-It should be easy so sample from the model. We just need $\mu$ and $\alpha(u_i, u_j)$ to specify the probabilities in \eqref{eq:latent}.  
-As the edge assignment in this latent model is independent of the other edges, we could just sample edges from a Bernoulli distribution with probabilities specified by \eqref{eq:latent}.
-However, I was not able to obtain $\mu$.
-So instead, to evaluate the fit, I followed the analysis in the book \cite{kolaczyk2014}.
-Here we fit a model to a subset of the network, and make the model predict the rest of the network.
-This was done using 5-fold cross-validation.
-
-A common evaluation metric is the ROC AUC, which is the area under the ROC curve of the predictions.
-Random assignment will results in AUC of approximately 0.5, while perfect classification will give AUC of 1.
-From our experiment we got a AUC of 0.54, so the model performs only slightly better than random edge assignment.
-This does not necessarily mean the model is a bad fit, it might just be that the network is hard to predict.
-For instance, edge prediction of a classical random graph would result is the same probability of all edges, and an AUC of $0.5$.
-
-\section{Summary}
-In this project we have looked at the social network of dolphins by \cite{dolphins}.
-We found that it was neither scale-free, nor had the small-world property.
-We could also not find any model that that was able to describe the social structure particularly well.
+To test for inclusion of
 

### 2016-11-25 project_code.R

@@ -138,7 +138,6 @@ survdiff(Surv(dager, dod)~bil_groups, data = pbc)
 survdiff(Surv(dager, dod)~alb_groups, data = pbc)
 
 #############################################################
-# b)
 # b) Univariat regresjon:
 # Deretter utfører du univariate Cox-regresjonsanalyser for hver av kovariatene. 
 # I denne sammenhengen ma du spesielt vurdere hvordan de numeriske kovariatene
@@ -195,7 +194,7 @@ dev.off()
 
 df = cox_res_uni(cox, 'Age')
 caption = 'Cox regression on the age covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
-label = 'tab:cox_treatment'
+label = 'tab:cox_age'
 tab =  xtable(df, caption = caption, label = label, 
               display = rep('g', 8), digits = 3)
 print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_age.tex')
@@ -204,21 +203,179 @@ print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_age.tex')
 # Small p-values indicate violation of proportionality assumption
 cox.zph(cox)
 plot(cox.zph(cox))
-
-?cox.zph
-?termplot
 termplot(cox, se=TRUE) # can be used when we fit spline in cox regression
 
-cox = coxph(Surv(dager, dod)~kjonn, data = pbc)
-?survfit
+#---------
+# Gender
+cox = coxph(Surv(dager, dod)~factor(kjonn), data = pbc)
+df = cox_res_uni(cox, 'Gender')
+caption = 'Cox regression on the gender covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+label = 'tab:cox_gender'
+tab =  xtable(df, caption = caption, label = label, 
+              display = rep('g', 8), digits = 3)
+print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_gender.tex')
+
+#-------
+# Bilirubin
+cox = coxph(Surv(dager, dod)~bil, data = pbc)
 cox
-cox = coxph(Surv(dager, dod)~bil_groups, data = pbc)
+martres = cox$residuals
+bil = pbc$bil
+printfig('cox_bil_mart', height = 4)
[…]

### 2016-11-29 latex/article.tex
more text

@@
 \todo[inline]{Introduce problem and covariates}
+This project is the first part of the final exam in STK4080/9080.
+
+Primary biliary cirrhosis, or PBC, is a rare liver decease. We do no currently know what cause this decease.  
+The American hospital Mayo followed 307 PBC patients in a double-blind study from January 1974 to May 1984, where the drug D-penicillamine was compared to placebo.
+Numerous covariates were recorded for the patients at the start of the studey, however, we will only work with a small subset of those.
+We have chosen to focus on the demographic variables gender and age, and the two biochemical covariates albumin (the dominating protein in blood plasma) and bilirubin (a yellow compompound that, in large quantities, is related with liver malfunction).
+
+\begin{itemize}
+    \item \verb+days+: Days from randomization to death or censoring. 
+    \item \verb++: 
+    \item \verb++: 
+    \item \verb++: 
+\end{itemize}
 
@@
 However, this does not affect the estimation of the Kaplan-Meier.
+For Cox regression however, ties can be a problem.
+There are different ways to handle ties here, but as there are very few ties compared to the size of the dataset, the different methods should yield more or less equivalent results.
+We therefore only use the Efron approximation in all our experiments.
 
 \section{a) Simple univariate analysis}
-\unsure[inline]{Does univariate mean only one group??? \\I would prefer to use many several groups.}
 \unsure[inline]{Should I try to interpret the K-M plots through quantiles and stuff?? I’m pretty sure I’ve seen that somewhere. Show median survival times for different groups?}
 
+To start with, we analyze the relationship between the death rates and the covariates, individually.
+
+Figure~\ref{fig:km_beh_ald},~\ref{fig:km_kjonn_bil}, and~\ref{fig:km_alb} shows the survival curves (Kaplan-Meier plots) for the individual covariates. 
+Under each plot we have added a histogram for the covariates. This is helpful for choosing the binning of numerical covariates.
+
+To the left in Figure~\ref{fig:km_beh_ald} we see that there seems to be very little difference between the treatment and placebo patients.
+There nothing here that suggest that treatment has a positive effect.
+Table~\ref{tab:log_rank_indiv} shows the log-rank test for the individual covariates, and we see here that treatment is not considered significant.
+
+To the right in Figure~\ref{fig:km_beh_ald} we see the survival curves for different age groups.
+There is quite a large difference between the oldest patients and the younger patients, suggesting that age play an important part in the death of the PBC patients.
+We also the that the age groups are considered significant by the log-rank test in Table~\ref{tab:log_rank_indiv}.
+However, I do not know how deaths are recorded in the dataset.
+If deaths cause by something not related to the PBC are recorded as deaths, it is possible that we only see the effect age has on humans in general.
+To truly say something about age, we should have a group of individuals without PBC and compare the death rates between the groups.
+%
 \begin{figure}[h!tbp]
@@
     \caption{Survival curves for treatment and age. We have included histograms for the two covariates under the survival curves.}
-    \label{fig:alb}
+    \label{fig:km_beh_ald}
 \end{figure}
 
+In Figure~\ref{fig:km_kjonn_bil}, we see the survival curves for gender and bilirubin.
+The figure might suggest that males has a higher hazard than females, but we see from the bar chart that there are quite few males in the dataset, so the difference might just be a results of small sample size.
+The log-rank test supports this hypothesis, as the $p$-value is quite small, but not very small.
+
+Bilirubin, however, seem to have a possibly large effect on the survival.
+Comparing the death ratios between the groups with the highest and lowest concentration of bilirubin, the difference is quite striking.
+Of course bilirubin is considered significant by the log-rank test.
+%
 \begin{figure}[h!tbp]
@@
     \caption{Survival curves for gender and amount of bilirubin in the blood of the test subjects. We have included histograms for the two covariates under the survival curves.}
-    \label{fig:alb}
+    \label{fig:km_kjonn_bil}
 \end{figure}
+
+Finally, we investigated the relationship between the concentration of albumin and the survival function in Figure~\ref{fig:km_alb}.
+Immediately, it is apparent that individuals with a higher concentration, also has higher death rates.
+The log-rank test also considers albumin as very significant.
 
@@
     \caption{Survival curve for amount of albumin in the blood of the test subjects. The histogram displays the distribution of albumin among the subjects.}
-    \label{fig:alb}
+    \label{fig:km_alb}
 \end{figure}
@@
 \input{tables/cox_bilirubin}
-\unsure[inline]{Something on how log transform is reasonable.}
+\unsure[inline]{Something on how log transform is reasonable. That differences are more important for small values of bilirubin than large.}
 
@@
 
+\info[inline]{Mention that we tested everything without log-transform, and it was worse}
+
 We see from Figure~\ref{fig:cor_heat} that the covariates are not particularly correlated.
@@
 And from a quick look at all the covariates plotted against each other, there does not seem to be any non-linear effect that cause the low correlations (figures not included).
+\change[inline]{Give English names to the covariates in the correlations figure}
 %
@@
 
-We started with all covariates and removed the ones that were not significant (just treatment).
+We started with all covariates and removed the ones that were not significant one by one. Thus we ended up with a mode containing age, log-transformed bilirubin, and albumin.
+After that, we added first order interaction to the models.
+This was first done for the model with age, log bilirubin, and albumin, where we only tested one interaction pair at a time (so first model as \verb$age * log(bil) + alb$).
+We then repeated the experiment with gender and treatment (one at a time).
+We did not use an ANOVA table to check the difference between the models, as all models except one had non-significant (on $5\%$ level) interactions.
+The only model with significant interaction was the gender-bilirubin interaction (at $p=0.023$.) However, this made the bilirubin covariate not significant.
+Investigating Figure~\ref{fig:km_kjonn_bil} we see that there are a lot more women than man in the dataset, which probably cause the gender-bilirubin interaction to just replace the bilirubin covariate.
+As a result we do not use this model further.
+\todo[inline]{Show code for this in appendix}
+
+\input{./tables/cox_final}
+
+\input{./tables/cox_final_ci}
+
+The score test of the final model has a $p$-value $< 1e-16$.
+
+\input{./tables/cox_time_dep}
+
+\input{./tables/cox_time_dep_log}
 
 To test for inclusion of
+
+\begin{figure}[h!tb]
+    \begin{center}
+        \includegraphics[scale=0.8]{./figures/cox_full_linearity.pdf}
+    \end{center}
+    \vspace{-0.8cm}
+    \caption{Martingale residuals for the full model.
+The lines are a GAM fitted to a smoothing spline of the relevant covariate.
+The bounds are the upper and lower pointwise twice-standard-error curves.}
+    \label{fig:cox_full_linearity}
+\end{figure}
 

### 2016-11-29 project_code.R

@@ -24,6 +24,8 @@ head(pbc)
 # We have tied data because of rounding to nearest day.
 length(pbc$dager) 
 length(unique(pbc$dager))
+sort(pbc$dager)
+?coxph
 #----------
 # a) Enkle univariate analyser:
 # Først utfører du enkle univariate analyser ved hjelp av Kaplan-Meier plot og 
@@ -253,6 +255,7 @@ plot(gam(martres~s(bil)), se=T, ylim=c(min(martres), max(martres)),
 points(bil, martres)
 abline(0, 0, lty=4)
 dev.off()
+
 df = cox_res_uni(cox, 'Bilirubin')
 caption = 'Cox regression on the log-transformed bilirubin covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
 label = 'tab:cox_bilirubin_log'
@@ -303,9 +306,11 @@ print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_albumin.tex')
 # (iii) kontrollere modellens forutsetninger.
 
 cox = coxph(Surv(dager, dod)~., data = pbc)
-cox = coxph(Surv(dager, dod)~factor(beh) + ald + factor(kjonn) + bil + alb, data = pbc)
+cox = coxph(Surv(dager, dod)~factor(beh) + ald + factor(kjonn) + log(bil) + alb, data = pbc)
 cox
-cox = coxph(Surv(dager, dod)~ald + factor(kjonn) + bil + alb, data = pbc)
+cox = coxph(Surv(dager, dod)~ald + factor(kjonn) + log(bil) + alb, data = pbc)
+cox
+cox = coxph(Surv(dager, dod)~ald + log(bil) + alb, data = pbc)
 cox
 
 library(lattice)
@@ -317,43 +322,68 @@ dev.off()
 plot(pbc$bil, pbc$alb, log = 'x', xlab = 'billirubin', ylab='albumin')
 plot(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')])
 
+# Test first-order combinations without gender and treatment
+cox = coxph(Surv(dager, dod)~ald * log(bil) + alb, data = pbc)
+cox #not
+cox = coxph(Surv(dager, dod)~ald * alb + log(bil), data = pbc)
+cox #not
+cox = coxph(Surv(dager, dod)~ald + log(bil) * alb, data = pbc)
+cox #not
+
[…]

### 2016-11-29 latex/article.tex
more text

@@
 Finally, we investigated the relationship between the concentration of albumin and the survival function in Figure~\ref{fig:km_alb}.
-Immediately, it is apparent that individuals with a higher concentration, also has higher death rates.
+Immediately, it is apparent that individuals with a lower concentration, has higher death rates.
+This is surprising, as a high concentration of albumin is usually connected to lowered functionality in the liver (see assignment text).
 The log-rank test also considers albumin as very significant.
@@
 
+Second, we analyzed the individual covariates by univariate Cox regression.
+As mentioned in the introduction, tied data poses a problem for Cox regression, but as we have few ties, the Efron approximation should be quite good.
+
 \subsection{Treatment}
 
-Here we fitted an cox regression to the treatment covariate.
-We see from Table~\ref{tab:log_rank_indiv} that the treatment and placebo group are not significantly different, and thus we do not expect the cox regression to be significant either.
+We start by fitting a cox regression to the treatment covariate.
+From the previous log-rank test in Table~\ref{tab:log_rank_indiv}, we found that the treatment and placebo groups are not significantly different, and thus we do not expect the cox regression to be significant either.
 This is because the score test in cox regression, with categorical covariates, is equivalent to the log-rank test, as long as there are no ties.
@@
 Table~\ref{tab:cox_treatment} shows some of the results from the cox regression.
-First note that we display the exponentiated coefficient, as this gives the hazard ratio between the groups, as well as the 90\% confidence interval.
+First note that we display the exponentiated coefficient, as this gives the hazard ratio between the groups, as well as the 95\% confidence interval.
 We see that the hazard rate is close to one, and 1 is covered in the confidence interval.
@@
 
-Also note that none of the tests for the proportional hazard assumption fails (zph).
-So the cox model seems to be a reasonable model for the treatment covariate.
+The columns \verb+zph km p+ and \verb+zph log p+ gives the $p$-value from the \verb+cox.zph+ function in R.
+This is a test for the proportional hazard assumption, using Schoenfeld-residuals.
+\verb+km+ and \verb+log+ denotes the Kaplan-Meier and log transform of the survival times.
+We see from the table than none of the tests suggest that the proportional hazard assumption is false.
 
@@
 
+Next, we fit a Cox regression to the age covariate, yielding the results in Table~\ref{tab:cox_age}.
+According to the score test, the age covariate is considered significant.
+The regression coefficient is quite small, but this is not surprising as the age is coded in years (in stead of e.g.\ decades).
+It suggest that the hazard ratio between two individuals with a age difference of one year is $1.04$.
+None of the zph tests suggest that the proportional hazard assumption is violated.
+
 \input{tables/cox_age}
 
-In Figure~\ref{fig:cox_age_mart} we have plotted the martingale residuals from the cox regression of age.
-There does not seem to be a pattern in the residuals, so the functional form of the covariate seems appropriate.
+In Figure~\ref{fig:cox_age_mart} we have plotted the martingale residuals from the cox regression.
+There does not seem to be a pattern in the residuals, so the functional form of the age covariate seems appropriate.
+The dense line in the figure shows a GAM fitted to a smoothing spline of the covariate, while the stippled lines give the upper and lower pointwise twice-standard-error curves.
 
@@
 \subsection{Gender}
+
+Next, we fitted a Cox regression to the gender covariate.
+As it is categorical, the score test in Table~\ref{tab:cox_gender} gives the same result as the log-rank test in Table~\ref{tab:log_rank_indiv}.
+In addition we see here that the zph tests suggest that the proportional hazard assumptions seem appropriate.
+
 \input{tables/cox_gender}
@@
 
-We fit a cox regression to the bilirubin covariate, and the results can be found in Table~\ref{tab:cox_bilirubin}, and the Martingale residuals in Figure~\ref{fig:cox_bil_mart}.
+Next, we fit a Cox regression to the bilirubin covariate.
+Results can be found in Table~\ref{tab:cox_bilirubin}, and the Martingale residuals in Figure~\ref{fig:cox_bil_mart}.
 The score test give a $p$-value of 0, but the we see a clear pattern in the martingale residuals.
-We therefore fit a cox regression with a cubic smoothing-spline to the bilirubin covariate, and got that both the linear and non-linear part was highly significant ($p < 1e-9$, table exempt).
+We therefore fit a Cox regression with a cubic smoothing-spline to the bilirubin covariate, and got that both the linear and non-linear part was highly significant ($p < 1e-9$, table exempt).
 
-The shape of the martingale residuals might suggest that a log-transform of the bilirubin variable could be a better fit.
+The shape of the martingale residuals might suggest that a log transform of the bilirubin variable could be a better fit.
 We also plot the spline term against the amount of bilirubin in Figure~\ref{fig:cox_bil_termplot}, where the x-axis is logarithmically scaled.
-We see that the line is almost linear, supporting our claim that a log-transform might be suitable.
+We see that the line is almost linear, supporting our claim that a log transform might be suitable.
 
-Next we fitted a cox regression to the log transformed bilirubin covariates, yielding the results in Table~\ref{tab:cox_bilirubin_log}.
-
-The martingale residuals were plotted in Figure~\ref{fig:cox_bil_mart_log}, where the residuals are evenly distributed around zero. This again support our log-transform.
+Next we fitted a Cox regression to the log transformed bilirubin covariates, yielding the results in Table~\ref{tab:cox_bilirubin_log}.
+The martingale residuals were plotted in Figure~\ref{fig:cox_bil_mart_log}, where the residuals are evenly distributed around zero. This again support our log transform.
+We see from the zph test that there is nothing suggesting that the proprtional hazard assumption is inappropriate.
 
 \input{tables/cox_bilirubin}
-\unsure[inline]{Something on how log transform is reasonable. That differences are more important for small values of bilirubin than large.}
 
@@
 \subsection{Albumin}
+
+Finally, we fitted a Cox regression to the albumin covariate. From Table~\ref{tab:cox_albumin} we see that the score test consider the albumin covariate as significant.
+According to the coefficient, we again see that a lower concentration of albumin is related to higher death rate.
+From the martingale residuals in Figure~\ref{fig:cox_alb_mart}, the functional form of the covariate seems appropriate. 
+However, the log transform of the proportional hazard test (\verb+zph log p+) is quite low, suggesting that there might be some time dependence in the hazards.
+There are different ways of handling this time dependence, but non of them were further pursued here.
+
 \input{tables/cox_albumin}
@@
 
-\subsection{Testing model assumptions}
-
-\begin{itemize}
-    \item Can plot martingale residuals vs time (observation number) and color by group. To see if the groups are evenly spread. This check the PH assumption.
-    \item Investigate martingale residuals. Figure this out!!!!
-    \item Scaled Schoenfeld residuals (lec 9 p 35) to check proportionality over time.
-    \item Schoenfeld residuals for prop. assumption (lec 9 p 38).
-    \item Check proportionality by lec 9 p 29: Plot log nelson aalen for different groups and check if they are parallel.
-    \item Practical exercise 8. Fit spline to check if log-linear effect (which is a model assumption?). How to interpret results?
-    \item \url{http://www.mwsug.org/proceedings/2006/stats/MWSUG-2006-SD08.pdf}
-\end{itemize}
-
 \section{c) Multivariate regression}
-\begin{itemize}
-    \item See Practical exercise 7 for likelihood ratio test for significance of interactions.
-\end{itemize}
-
 \info[inline]{Mention that we tested everything without log-transform, and it was worse}
 
-We see from Figure~\ref{fig:cor_heat} that the covariates are not particularly correlated.
+The last part of this assignment was to fit a multivariate Cox regression to the covariates, and find a model that describes the survival.
+
+As a first step, we found the correlations between the covariates.
+From Figure~\ref{fig:cor_heat} we see that the covariates are not particularly correlated.
 The highest (in absolute value) is between bilirubin and albumin on approximately $-0.4$.
 And from a quick look at all the covariates plotted against each other, there does not seem to be any non-linear effect that cause the low correlations (figures not included).
-\change[inline]{Give English names to the covariates in the correlations figure}
+
+In this section we only work with the log transformed bilirubin covariate, as we found that is was an appropriate transform in the previous section.
 %
@@
 
-We started with all covariates and removed the ones that were not significant one by one. Thus we ended up with a mode containing age, log-transformed bilirubin, and albumin.
+To find an appropriate set of covariates for the Cox model, we started with all covariates and removed the ones that were not significant one by one.
+Thus we ended up with a model containing age, bilirubin, and albumin.
 After that, we added first order interaction to the models.
-This was first done for the model with age, log bilirubin, and albumin, where we only tested one interaction pair at a time (so first model as \verb$age * log(bil) + alb$).
-We then repeated the experiment with gender and treatment (one at a time).
+This was first done for the model with age, bilirubin, and albumin, where we only tested one interaction pair at a time.
+We then repeated this interaction experiment with gender and treatment (one at a time).
 We did not use an ANOVA table to check the difference between the models, as all models except one had non-significant (on $5\%$ level) interactions.
@@
 
+Our final model therefore contains age, bilirubin, and albumin, without any interactions. We therefore conclude that there is no evidence that the treatment has an effect on the patients.
+
+\todo[inline]{Mention again that age could just be age\ldots}
 \input{./tables/cox_final}

### 2016-11-29 project_code.R

@@ -18,7 +18,7 @@ head(pbc)
 # plot(pbc$beh, pbc$dod)
 # plot(pbc)
 # plot(pbc$kjonn, pbc$ald)
- 
+?cox.zph 
 
 # Not unique death / censor times
 # We have tied data because of rounding to nearest day.
@@ -293,7 +293,6 @@ label = 'tab:cox_albumin'
 tab =  xtable(df, caption = caption, label = label, 
               display = rep('g', 8), digits = 3)
 print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_albumin.tex')
-
 
 #############################################################
 # c) Multivariat regresjon:

### 2016-11-29 latex/article.tex
more text

@@
 Our final model therefore contains age, bilirubin, and albumin, without any interactions. We therefore conclude that there is no evidence that the treatment has an effect on the patients.
+\todo[inline]{Mention again that age could just be age\ldots}
 
-\todo[inline]{Mention again that age could just be age\ldots}
+In Table~\ref{tab:cox_final} and~\ref{tab:cox_final_ci} we show the resuting coefficients with confidence intervals and $p$-values.
+We see that the coefficient for age has not changes from the univariate case, while the coefficients for bilirubin and albumin has changed more.
+This probably because there is a stronger dependence between bilirubin and albumin.
+
+\todo[inline]{Interpret coefficients}
+
 \input{./tables/cox_final}
@@
 
+In Table~\ref{tab:cox_time_dep} and~\ref{tab:cox_time_dep_log} we have tested the proportional hazard assumption of the model, using Schoenfeld-residuals.
+We see that non of the test indicate that there is a time dependency in the hazards.
+However, we find lowest $p$-values for the albumin covariate, so it could be interesting to further investigate this.
+This was, however, not pursued in this project.
+
 \input{./tables/cox_time_dep}
@@
 
-To test for inclusion of
+In Figure~\ref{fig:cox_full_linearity}, we have plotted the martingale residuals against the individual covariates.
+The residual seem to be evenly distributed around zero, suggesting that the functional form of our covariates are appropriate.
 

### 2016-11-29 project_code.R

@@ -314,9 +314,14 @@ cox
 
 library(lattice)
 printfig('cor_heat', height = 5, width = 5)
-levelplot(cor(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')]), pretty=TRUE, xlab='',
-          ylab='')
+x = cor(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')])
+row.names(x) = c('treat', 'age', 'gender', 'bil', 'alb')
+colnames(x) = row.names(x)
+levelplot(x, pretty=TRUE, xlab='', ylab='')
+#levelplot(cor(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')]), pretty=TRUE, xlab='',
+#          ylab='')
 dev.off()
+?levelplot
 
 plot(pbc$bil, pbc$alb, log = 'x', xlab = 'billirubin', ylab='albumin')
 plot(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')])
@@ -405,7 +410,7 @@ cox.zph(cox, transform = log)
 # Interpret model
 cs = summary(cox)
 cs
-df = data.frame(cs$coefficients)
+df = data.frame(cs$coefficients, check.names = FALSE)
 row.names(df) = c('age', 'log(bilirubin)', 'albimin')
 caption = 'Coefficinets for cox regression.'
 label = 'tab:cox_final'
@@ -415,7 +420,7 @@ print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_final.tex')
 
 pars = cs$conf.int
 pars
-df = data.frame(pars)
+df = data.frame(pars, check.names = FALSE)
 row.names(df) = c('age', 'log(bilirubin)', 'albimin')
 caption = 'CI for cox regression.'
 label = 'tab:cox_final_ci'
@@ -425,18 +430,18 @@ print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_final_ci.tex')
 
 cs$sctest
 zph = cox.zph(cox)
-df = data.frame(zph$table)
+df = data.frame(zph$table, check.names = FALSE)
 row.names(df) = c('age', 'log(bilirubin)', 'albimin', 'GLOBAL')
-caption = 'Test for time dependancy.'
+caption = 'Test for proportional hazard using Schoenfeld-residuals. K-M transfromed.'
 label = 'tab:cox_time_dep'
 tab =  xtable(df, caption = caption, label = label, 
               display = rep('g', 4), digits = 3)
[…]

### 2016-11-30 latex/article.tex
more text

--- 2016-11-29 latex/article.tex
@@
 \section{Introduction}
-\todo[inline]{Introduce problem and covariates}
+\todo[inline]{Add code}
 This project is the first part of the final exam in STK4080/9080.
 
-Primary biliary cirrhosis, or PBC, is a rare liver decease. We do no currently know what cause this decease.  
+Primary biliary cirrhosis, or PBC, is a rare liver decease. We do not currently know what cause this decease.  
 The American hospital Mayo followed 307 PBC patients in a double-blind study from January 1974 to May 1984, where the drug D-penicillamine was compared to placebo.
-Numerous covariates were recorded for the patients at the start of the studey, however, we will only work with a small subset of those.
-We have chosen to focus on the demographic variables gender and age, and the two biochemical covariates albumin (the dominating protein in blood plasma) and bilirubin (a yellow compompound that, in large quantities, is related with liver malfunction).
-
-\begin{itemize}
-    \item \verb+days+: Days from randomization to death or censoring. 
-    \item \verb++: 
-    \item \verb++: 
-    \item \verb++: 
-\end{itemize}
-
+Numerous covariates were recorded for the patients at the start of the study, of which we will only work with a small subset.
+We have chosen to focus on the demographic variables gender and age, and the two biochemical covariates albumin (the dominating protein in blood plasma) and bilirubin (a yellow compound that, in large quantities, is related with liver malfunction).
+\\
+\\
 The dataset has tied events due to the rounding of events to nearest date.
-Out of the 307 observations, 296 occurs alone in a single date.
-However, this does not affect the estimation of the Kaplan-Meier.
+Out of the 307 observations, there are 296 different days of occurrences.
+However, this does not affect the estimation of Kaplan-Meier.
 For Cox regression however, ties can be a problem.
-There are different ways to handle ties here, but as there are very few ties compared to the size of the dataset, the different methods should yield more or less equivalent results.
+There are different ways to handle ties here, but as there are very few ties compared to the size of the dataset, the methods should yield more or less equivalent results.
 We therefore only use the Efron approximation in all our experiments.
@@
 \section{a) Simple univariate analysis}
-\unsure[inline]{Should I try to interpret the K-M plots through quantiles and stuff?? I’m pretty sure I’ve seen that somewhere. Show median survival times for different groups?}
 
@@
 Figure~\ref{fig:km_beh_ald},~\ref{fig:km_kjonn_bil}, and~\ref{fig:km_alb} shows the survival curves (Kaplan-Meier plots) for the individual covariates. 
-Under each plot we have added a histogram for the covariates. This is helpful for choosing the binning of numerical covariates.
+Under each plot we have added a histogram for the covariates. This is helpful for choosing the grouping of numerical covariates.
 
 To the left in Figure~\ref{fig:km_beh_ald} we see that there seems to be very little difference between the treatment and placebo patients.
-There nothing here that suggest that treatment has a positive effect.
+There is nothing here that suggest that treatment has a positive effect.
 Table~\ref{tab:log_rank_indiv} shows the log-rank test for the individual covariates, and we see here that treatment is not considered significant.
@@
 There is quite a large difference between the oldest patients and the younger patients, suggesting that age play an important part in the death of the PBC patients.
-We also the that the age groups are considered significant by the log-rank test in Table~\ref{tab:log_rank_indiv}.
+We also see that the age groups are considered significant by the log-rank test in Table~\ref{tab:log_rank_indiv}.
 However, I do not know how deaths are recorded in the dataset.
-If deaths cause by something not related to the PBC are recorded as deaths, it is possible that we only see the effect age has on humans in general.
+If deaths caused by something not related to PBC are recorded as deaths, it is possible that we only see the effect age has on humans in general.
 To truly say something about age, we should have a group of individuals without PBC and compare the death rates between the groups.
@@
 Finally, we investigated the relationship between the concentration of albumin and the survival function in Figure~\ref{fig:km_alb}.
-Immediately, it is apparent that individuals with a lower concentration, has higher death rates.
+Immediately, it is apparent that individuals with lower concentrations, have higher death rates.
 This is surprising, as a high concentration of albumin is usually connected to lowered functionality in the liver (see assignment text).
@@
 \section{b) Univariate regression}
-\info[inline]{For categorical covariates, cox scores and log-rank are equivalent, if not ties. What with ties??}
-\unsure[inline]{How does cox handle tied data?}
-\info[inline]{Cox can fail if $x$ is wrong and we should use log or sqrt instead. Or if the hazards are not proportional.}
 
 Second, we analyzed the individual covariates by univariate Cox regression.
-As mentioned in the introduction, tied data poses a problem for Cox regression, but as we have few ties, the Efron approximation should be quite good.
+As mentioned in the introduction, tied data poses a problem for Cox regression, but as we have few ties, the Efron approximation should work well.
 
@@
 Table~\ref{tab:cox_treatment} shows some of the results from the cox regression.
-First note that we display the exponentiated coefficient, as this gives the hazard ratio between the groups, as well as the 95\% confidence interval.
+First note that we display the exponentiated coefficient, as this gives the hazard ratio between the groups.
+We also display the 95\% confidence interval.
 We see that the hazard rate is close to one, and 1 is covered in the confidence interval.
@@
 \verb+km+ and \verb+log+ denotes the Kaplan-Meier and log transform of the survival times.
-We see from the table than none of the tests suggest that the proportional hazard assumption is false.
+We see from the table that none of the tests suggest that the proportional hazard assumption is false.
 
@@
 Results can be found in Table~\ref{tab:cox_bilirubin}, and the Martingale residuals in Figure~\ref{fig:cox_bil_mart}.
-The score test give a $p$-value of 0, but the we see a clear pattern in the martingale residuals.
+The score test give a $p$-value of 0, but we see a clear pattern in the martingale residuals, suggesting that the functional form is not correct.
 We therefore fit a Cox regression with a cubic smoothing-spline to the bilirubin covariate, and got that both the linear and non-linear part was highly significant ($p < 1e-9$, table exempt).
+This suggest that the covariate should be transformed.
 
@@
 \section{c) Multivariate regression}
-\info[inline]{Mention that we tested everything without log-transform, and it was worse}
 
 The last part of this assignment was to fit a multivariate Cox regression to the covariates, and find a model that describes the survival.
-
 As a first step, we found the correlations between the covariates.
@@
 
-To find an appropriate set of covariates for the Cox model, we started with all covariates and removed the ones that were not significant one by one.
-Thus we ended up with a model containing age, bilirubin, and albumin.
+To find an appropriate set of covariates for the Cox model, we started with all covariates and removed the ones that were not significant, one at a time.
+Thus, we ended up with a model containing age, bilirubin, and albumin.
 After that, we added first order interaction to the models.
@@
 Our final model therefore contains age, bilirubin, and albumin, without any interactions. We therefore conclude that there is no evidence that the treatment has an effect on the patients.
+\todo[inline]{Mention again that age could just be age\ldots}
 
-\todo[inline]{Mention again that age could just be age\ldots}
+In Table~\ref{tab:cox_final} and~\ref{tab:cox_final_ci} we show the resulting coefficients with confidence intervals and $p$-values.
+We see that the coefficient for age has not changes from the univariate case, while the coefficients for bilirubin and albumin have changed more.
+This probably because there is a stronger dependence between bilirubin and albumin.
+
+Coefficients ($\beta_i$) in a Cox regression are typically interpret in the following way.
+An increase of one unit in a covariate (give all other covariates fixed) corresponds to $\exp(\beta_i)$ ratio between the two hazards.
+For the age covariate, this means that the hazard ratio of two individuals with one year difference is $1.04$, or an increased hazard of $4\%$ for a patient that is one year older than an other (identical except age) patient.
+As the hazard increases exponentially with age, this is quite a large number.
+It could have been more fitting to use $\text{age}/ 10$, as the coefficient would possibly be more interpretable.
+
+Albumin also has an exponential relationship with the hazard, though the hazard decrease with the concentration of albumin. For an increase of one mg/dl albumin the hazard decrease with $32\%$.
+
+For bilirubin, we have used the log transform of the covariate instead of the covariate.
+This makes the interpretation a bit different.
+For one unit increase in the logarithm of bilirubin, we have an increase of $271\%$.
+However, as the coefficient is quite close to one, the hazard is approximately proportional to bilirubin. So a doubling of the concentration of bilirubin, corresponds to a doubling of the hazard.
+
 \input{./tables/cox_final}
@@
 
-The score test of the final model has a $p$-value $< 1e-16$.
+In Table~\ref{tab:cox_time_dep} and~\ref{tab:cox_time_dep_log} we have tested the proportional hazard assumption of the model, using Schoenfeld-residuals.
+We see that non of the tests indicate that there is a time dependency in the hazards.
+However, we find the smallest $p$-values for albumin, so it could be interesting to further investigate this.
+That was, however, not pursued in this project.
 
@@
 
+\cleardoublepage{}
+\begin{appendices}
+    \section{a) Simple univariate analysis}
+    \begin{verbatim}
+    pbc <- read.csv("~/stk4080/pbc3A.txt", sep="")
+
+    ## Treatment
+    layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
+    fit = survfit(Surv(dager, dod)~beh, data = pbc)
+    survfit(Surv(pbc$dager, pbc$dod)~pbc$beh)
+    plot(fit, lty=1:2, ylab='survival', xlab='days')
+    legend(1,0.2,c("treatment","placebo"), lty=1:2, bty='n')
+    grid()
+    plot(as.factor(pbc$beh), main='', ylab='frequency', xlab='treatment', col='white',
+         names.arg=c('treatment', 'placebo'))
+
+    ## Age
+    layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
+    pbc$ald_groups = 0
+    gr = c(0, 40, 50, 60)
+    for (i in gr) {
+      pbc$ald_groups[pbc$ald > i] = i
+      }
+    fit = survfit(Surv(dager, dod)~ald_groups, data = pbc)
+    plot(fit, lty=1:length(gr), ylab='survival', xlab='days')
+    legend(1,0.3, str_lab(gr), lty=1:length(gr), bty='n')
+    grid()
+    hist(pbc$ald, breaks = 20, main='', ylab='frequency', xlab='age')
+
+    ## Gender
+    layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
+    fit = survfit(Surv(dager, dod)~kjonn, data = pbc)
+    plot(fit, lty=1:2, ylab='survival', xlab='days')
+    legend(1,0.2,c("male","female"), lty=1:2, bty='n')
+    grid()
+    plot(as.factor(pbc$kjonn), main='', ylab='frequency', xlab='gender', col='white',
+         names.arg=c('male', 'female'))
+
+    ## Bilirubin
+    layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
+    pbc$bil_groups = 0
+    gr = c(0, 1, 3, 7, 15)
+    for (i in gr) {
+      pbc$bil_groups[pbc$bil > i] = i
+      }
+    fit = survfit(Surv(dager, dod)~bil_groups, data = pbc)
+    plot(fit, lty=1:length(gr), ylab='survival', xlab='days')
+    legend(1,0.4, str_lab(gr), lty=1:length(gr), bty='n')
+    grid()
+    hist(pbc$bil, breaks = 20, main='', ylab='frequency', xlab='bilirubin (mg/dl)')
+
+    ## Albumin
+    layout(matrix(c(rep(1, 3), rep(2, 2)), nrow = 5, ncol = 1))
+    pbc$alb_groups = 0
+    gr = c(0, 3, 3.5, 4)
+    for (i in gr) {
+      pbc$alb_groups[pbc$alb > i] = i
+      }
+    fit = survfit(Surv(dager, dod)~alb_groups, data = pbc)
+    plot(fit, lty=1:length(gr), ylab='survival', xlab='days')
+    legend(1, 0.4, str_lab(gr), lty=1:length(gr), bty='n')
+    grid()
+
+      hist(pbc$alb, breaks = 20, main = '', ylab = 'frequency', xlab = 'albumin (mg/dl)')
+
+    ## Log-rank tests
+    survdiff(Surv(dager, dod)~beh, data = pbc)
+    survdiff(Surv(dager, dod)~ald_groups, data = pbc)
+    survdiff(Surv(dager, dod)~kjonn, data = pbc)
+    survdiff(Surv(dager, dod)~bil_groups, data = pbc)
+    survdiff(Surv(dager, dod)~alb_groups, data = pbc)
+
+    \end{verbatim}
+
+    \section{b) Univariate regression}
+    \begin{verbatim}
+
+    cox_res_uni = function(cox, cov_name){
+      cs = summary(cox)
+      pars = cs$conf.int
+      sctest = cs$sctest[3]
+      zph = cox.zph(cox)$table[3]
+      zph_log = cox.zph(cox, transform = log)$table[3]
+
+      df = data.frame(matrix(c(pars, sctest, zph, zph_log), nrow=1))
+      names(df) = c(colnames(pars), 'score p', 'zph km p', 'zph log p')
+      row.names(df) = cov_name
+      df
+    }
+
+    ## Treatment
+    cox = coxph(Surv(dager, dod)~beh, data = pbc)
+    df = cox_res_uni(cox, 'Treatment')
+    caption = 'Cox regression on treatment covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+    label = 'tab:cox_treatment'
+    tab =  xtable(df, caption = caption, label = label, 
+                  display = rep('g', 8), digits = 3)
+    print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_treatment.tex')
+
+    ## Age
+    cox = coxph(Surv(dager, dod)~ald, data = pbc)
+    martres = cox$residuals
+    plot(pbc$ald, martres)
+    ald = pbc$ald
+    plot(gam(martres~s(ald)), se=T, ylim=c(min(martres), max(martres)), 
+        xlab='age', ylab='martingale residual')
+    points(ald, martres)
+    abline(0, 0, lty=4)
+
+    df = cox_res_uni(cox, 'Age')
+    caption = 'Cox regression on the age covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+    label = 'tab:cox_age'
+    tab =  xtable(df, caption = caption, label = label, 
+              display = rep('g', 8), digits = 3)
+              print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_age.tex')
+
+    ## Gender
+    cox = coxph(Surv(dager, dod)~factor(kjonn), data = pbc)
+    df = cox_res_uni(cox, 'Gender')
+    caption = 'Cox regression on the gender covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+    label = 'tab:cox_gender'
+    tab =  xtable(df, caption = caption, label = label, 
+              display = rep('g', 8), digits = 3)
+              print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_gender.tex')
+
+    ## Bilirubin
+    cox = coxph(Surv(dager, dod)~bil, data = pbc)
+    martres = cox$residuals
+    bil = pbc$bil
+    plot(gam(martres~s(bil)), se=T, ylim=c(min(martres), max(martres)), 
+     xlab='bilirubin', ylab='martingale residual')
+     points(bil, martres)
+     abline(0, 0, lty=4)
+
+     df = cox_res_uni(cox, 'Bilirubin')
+     caption = 'Cox regression on the bilirubin covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+     label = 'tab:cox_bilirubin'
+     tab =  xtable(df, caption = caption, label = label, 
+              display = rep('g', 8), digits = 3)
+              print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_bilirubin.tex')
+
+    cox = coxph(Surv(dager, dod)~pspline(bil), data = pbc)
+    termplot(cox, se=TRUE, log='x', xlab='bilirubin', 
+    ylab='partial for pspline(bilirubin)')
+
+    cox = coxph(Surv(dager, dod)~log(bil), data = pbc)
+    martres = cox$residuals
+    bil = pbc$bil
+    plot(gam(martres~s(bil)), se=T, ylim=c(min(martres), max(martres)), 
+        xlab='bilirubin', ylab='martingale residual', log='')
+    points(bil, martres)
+    abline(0, 0, lty=4)
+
+    df = cox_res_uni(cox, 'Bilirubin')
+    caption = 'Cox regression on the log-transformed bilirubin covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+    label = 'tab:cox_bilirubin_log'
+    tab =  xtable(df, caption = caption, label = label, 
+              display = rep('g', 8), digits = 3)
+    print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_bilirubin_log.tex')
+
+    ## Albumin
+    cox = coxph(Surv(dager, dod)~alb, data = pbc)
+    martres = cox$residuals
+    alb = pbc$alb
+    plot(gam(martres~s(alb)), se=T, ylim=c(min(martres), max(martres)), 
+         xlab='albumin', ylab='martingale residual')
+    points(alb, martres)
+    abline(0, 0, lty=4)
+
+    df = cox_res_uni(cox, 'Albumin')
+    caption = 'Cox regression on the albumin covariate. score is the score test, zph is the proportional hazard test using scaled Schoenfeld-residuals.'
+    label = 'tab:cox_albumin'
+    tab =  xtable(df, caption = caption, label = label, 
+                  display = rep('g', 8), digits = 3)
+    print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_albumin.tex')
+
+    \end{verbatim}
+
+    \section{c) Multivariate regression}
+    \begin{verbatim}
+    \end{verbatim}
+
+\end{appendices}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\cleardoublepage{}
-\phantomsection
-\addcontentsline{toc}{chapter}{Bibliography}
-%\bibliographystyle{plain}
-%\bibliographystyle{apa}
-%\bibliographystyle{authordate4}
-%\bibliographystyle{ksfh_nat}
-\bibliographystyle{plainnat}
-\bibliography{bibliography}
+%\cleardoublepage{}
+%\phantomsection
+%\addcontentsline{toc}{chapter}{Bibliography}
+%%\bibliographystyle{plain}
+%%\bibliographystyle{apa}
+%%\bibliographystyle{authordate4}
+%%\bibliographystyle{ksfh_nat}
+%\bibliographystyle{plainnat}
+%\bibliography{bibliography}
 

### 2016-11-30 project_code.R

@@ -449,7 +449,6 @@ print(tab, type = 'latex',  file = '~/stk4080/latex/tables/cox_time_dep_log.tex'
 
 
 
-
 cox = coxph(Surv(dager, dod)~pspline(ald) + factor(kjonn) + pspline(bil) + pspline(alb), data = pbc)
 cox 
 cox = coxph(Surv(dager, dod)~ald + factor(kjonn) + pspline(bil) + alb, data = pbc)

### 2016-12-01 latex/article.tex
finished

@@
 \section{Introduction}
-\todo[inline]{Add code}
 This project is the first part of the final exam in STK4080/9080.
@@
 As a result we do not use this model further.
-\todo[inline]{Show code for this in appendix}
 
 Our final model therefore contains age, bilirubin, and albumin, without any interactions. We therefore conclude that there is no evidence that the treatment has an effect on the patients.
-\todo[inline]{Mention again that age could just be age\ldots}
 
@@
     \begin{verbatim}
+
+    ## Correlation plot
+    x = cor(pbc[c('beh', 'ald', 'kjonn', 'bil', 'alb')])
+    row.names(x) = c('treat', 'age', 'gender', 'bil', 'alb')
+    colnames(x) = row.names(x)
+    levelplot(x, pretty=TRUE, xlab='', ylab='')
+
+    cox = coxph(Surv(dager, dod)~factor(beh) + ald + factor(kjonn) + log(bil) + alb,
+                data = pbc)
+    cox
+    cox = coxph(Surv(dager, dod)~ald + factor(kjonn) + log(bil) + alb, data = pbc)
+    cox
+    cox = coxph(Surv(dager, dod)~ald + log(bil) + alb, data = pbc)
+    cox
+
+    ## Test first-order combinations without gender and treatment
+    cox = coxph(Surv(dager, dod)~ald * log(bil) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald * alb + log(bil), data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald + log(bil) * alb, data = pbc)
+    cox
+
+    ## Test first-order interactions without treatment
+    cox = coxph(Surv(dager, dod)~ald*factor(kjonn) + log(bil) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald*log(bil) + factor(kjonn) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald*alb + factor(kjonn) + log(bil), data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald + factor(kjonn)*log(bil) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald + factor(kjonn)*alb + log(bil), data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald + factor(kjonn) + log(bil)*alb, data = pbc)
+    cox 
+    # Investigate the interesting without gender 
+    cox = coxph(Surv(dager, dod)~ald*log(bil) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~ald + (kjonn)*log(bil) + alb, data = pbc)
+    cox
+
+    ## Test interactions with treatment
+    cox = coxph(Surv(dager, dod)~factor(beh)*ald + factor(kjonn) + log(bil) + alb, 
+        data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~factor(beh)*factor(kjonn) + ald + log(bil) + alb, 
+        data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~factor(beh)*log(bil) + ald + factor(kjonn) + alb,
+        data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~factor(beh)*alb + ald + factor(kjonn) + log(bil),
+        data = pbc)
+    cox
+
+    cox = coxph(Surv(dager, dod)~factor(beh)*ald + log(bil) + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~factor(beh)*log(bil) + ald + alb, data = pbc)
+    cox 
+    cox = coxph(Surv(dager, dod)~factor(beh)*alb + ald + log(bil), data = pbc)
+    cox
+
     \end{verbatim}
